---
title: "HMM_modeling"
author: "Shinnosuke Yagi"
date: "2025-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import libraries

```{r}
library(ape)
library(spam)
library(usethis)
library(devtools)
library(INLA)
library(phytools)
library(devtools)
library(phylodyn)
library(knitr)
library(ggplot2)
library(stringr)
library(jsonlite)
library(dplyr)
library(reticulate)
library(reshape2)
library(tidyr)
```

## Add "state" column to the dataframe, indicating whether the tree position is under selection (1) or neutral (0)

```{r}
all_files <- list.files("/Users/yagishinnosuke/Documents/2024-2025 Stanford/Research/Selective-Sweep-Simulation/Results/Two_Sample_Test_ARG_0.1/CSVs_run2", pattern = "\\.csv$", full.names = TRUE)

# Parameters
selected_site <- 25000  # Location of selection in bp

for (file in all_files) {
  df <- read.csv(file)
  df <- df %>%
    mutate(state = ifelse(left <= selected_site & right > selected_site, 1, 0))
  
  # Overwrite the same file or save a new one
  write.csv(df, file, row.names = FALSE)
}
```

## Generate F-matrix for all the newick trees

```{r}
all_chromosomes <- list()

for (file_index in seq_along(all_files)) {
  file <- all_files[file_index]
  df <- read.csv(file)

  message(sprintf("Processing file %d of %d: %s", file_index, length(all_files), basename(file)))
  
  chrom_trees <- list()
  
  for (i in seq_len(nrow(df))) {
    newick_str <- df$newick[i]
    state_label <- df$state[i]
    left_pos <- df$left[i]
    right_pos <- df$right[i]

    if (is.na(newick_str) || newick_str == "") next

    tryCatch({
      tree <- read.tree(text = newick_str)
      Fmat <- gen_Fmat(tree)

      chrom_trees[[length(chrom_trees) + 1]] <- list(
        Fmat = Fmat,
        state = state_label,
        left = left_pos,
        right = right_pos
      )

      counter <- counter + 1
      if (counter %% 100 == 0) {
        # message(sprintf("...Processed %d trees so far", counter))
      }
    }, error = function(e) {
     #  warning(sprintf("Skipping row %d in file %s: %s", i, basename(file), e$message))
    })
  }

  # Store each chromosome's list of trees
  chrom_name <- tools::file_path_sans_ext(basename(file))
  all_chromosomes[[chrom_name]] <- chrom_trees
}
```

```{r}
saveRDS(all_chromosomes, "/Users/yagishinnosuke/Documents/2024-2025 Stanford/Research/Selective-Sweep-Simulation/Results/Two_Sample_Test_ARG_0.1/all_chromosomes_grouped.rds")
```

```{r}
all_trees <- list()
counter <- 0

for (file_index in seq_along(all_files)) {
  file <- all_files[file_index]
  df <- read.csv(file)

  message(sprintf("Processing file %d of %d: %s", file_index, length(all_files), basename(file)))

  for (i in seq_len(nrow(df))) {
    newick_str <- df$newick[i]
    state_label <- df$state[i]
    left_pos <- df$left[i]
    right_pos <- df$right[i]

    if (is.na(newick_str) || newick_str == "") next

    tryCatch({
      tree <- read.tree(text = newick_str)
      Fmat <- gen_Fmat(tree)

      all_trees[[length(all_trees) + 1]] <- list(
        Fmat = Fmat,
        state = state_label,
        left = left_pos,
        right = right_pos
      )

      counter <- counter + 1
      if (counter %% 100 == 0) {
        message(sprintf("...Processed %d trees so far", counter))
      }
    }, error = function(e) {
      warning(sprintf("Skipping row %d in file %s: %s", i, basename(file), e$message))
    })
  }
}

# Save with genomic positions now included
saveRDS(
  all_trees,
  file = "/Users/yagishinnosuke/Documents/2024-2025 Stanford/Research/Selective-Sweep-Simulation/Results/Two_Sample_Test_ARG_0.1/all_trees_with_Fmat.rds"
)

message(sprintf("✅ Done! Total trees processed: %d", counter))
```

## Split data into training and test set

```{r}
# Get all chromosome names
set.seed(42)
chrom_names <- names(all_chromosomes)
n_total <- length(chrom_names)

# Split chromosome-level
n_train <- floor(0.6 * n_total)
n_val <- floor(0.3 * n_total)
n_test <- n_total - n_train - n_val

train_chroms <- chrom_names[1:n_train]
val_chroms <- chrom_names[(n_train + 1):(n_train + n_val)]
test_chroms <- chrom_names[(n_train + n_val + 1):n_total]

train_set <- all_chromosomes[train_chroms]
val_set <- all_chromosomes[val_chroms]
test_set <- all_chromosomes[test_chroms]

rm(all_chromosomes)
gc()
```

```{r}
# Split data into train (60%), validation (30%), test (10%)
set.seed(42)
n <- length(tree_data)
indices <- sample(seq_len(n))

n_train <- floor(0.6 * n)
n_val   <- floor(0.3 * n)
n_test  <- n - n_train - n_val

train_set <- tree_data[indices[1:n_train]]
val_set   <- tree_data[indices[(n_train + 1):(n_train + n_val)]]
test_set  <- tree_data[indices[(n_train + n_val + 1):n]]
```

## Compute $F_0$ and $F_1$

```{r}
# Function to pad F-matrices to a target size
pad_Fmat <- function(Fmat, target_dim) {
  m <- nrow(Fmat)
  if (m == target_dim) return(Fmat)
  padded <- matrix(0, nrow = target_dim, ncol = target_dim)
  padded[1:m, 1:m] <- Fmat
  return(padded)
}

# Filter F_1 from training set (state == 1)
# F1_list <- lapply(train_set, function(x) {
#   if (x$state == 1) x$Fmat else NULL
# })
F1_list <- lapply(unlist(train_set, recursive = FALSE), function(x) {
  if (x$state == 1) x$Fmat else NULL
})
F1_list <- Filter(Negate(is.null), F1_list)

# Filter F_0 from training set (state == 0 AND far from selected site)
# F0_list <- lapply(train_set, function(x) {
#   if (x$state == 0 && !is.null(x$left) && x$left > 50000) x$Fmat else NULL
# })
F0_list <- lapply(unlist(train_set, recursive = FALSE), function(x) {
  if (x$state == 0 && !is.null(x$left) && x$left > 50000) x$Fmat else NULL
})
F0_list <- Filter(Negate(is.null), F0_list)

# Pad all to the max dimension
max_dim <- max(sapply(c(F0_list, F1_list), function(F) nrow(F)))
F0_list_padded <- lapply(F0_list, pad_Fmat, target_dim = max_dim)
F1_list_padded <- lapply(F1_list, pad_Fmat, target_dim = max_dim)

# Average function
average_Fmat <- function(F_list) {
  Reduce("+", F_list) / length(F_list)
}

# Compute average F-matrices
F_0 <- average_Fmat(F0_list_padded)
F_1 <- average_Fmat(F1_list_padded)

# Optionally save
saveRDS(list(train = train_set, val = val_set, test = test_set), "/Users/yagishinnosuke/Documents/2024-2025 Stanford/Research/Selective-Sweep-Simulation/Results/tree_data_split.rds")
saveRDS(list(F_0 = F_0, F_1 = F_1), "/Users/yagishinnosuke/Documents/2024-2025 Stanford/Research/Selective-Sweep-Simulation/Results/F0_F1_from_training.rds")
```

## Visualize $F_0$ and $F_1$ matrices

```{r}
# Helper: plot an F-matrix
plot_F_matrix <- function(Fmat, title = "F-matrix") {
  # Convert to data frame and check dimensions
  if (!is.matrix(Fmat)) stop("Input must be a matrix")
  F_df <- as.data.frame(Fmat)
  F_df$i <- nrow(F_df):1  # for y-axis reversal
  F_long <- reshape2::melt(F_df, id.vars = "i")
  colnames(F_long) <- c("i", "j", "value")

  ggplot(F_long, aes(x = j, y = i, fill = value)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(title = title, x = "j", y = "i") +
    theme_minimal() +
    coord_fixed()
}

# Example usage (assuming F_0 and F_1 are defined in your environment)
plot_F_matrix(F_0, "Average F-matrix (Neutral, F_0)")
plot_F_matrix(F_1, "Average F-matrix (Selective, F_1)")
```

## Compute emission probabilities for F-matrices

```{r}
# Input: 
#   F_list: list of F-matrices (test set)
#   F_0, F_1: average F-matrices from training
#   beta: emission sharpness parameter
# Output:
#   Emission matrix E (T x 2)

compute_emission_matrix <- function(F_list, F_0, F_1, beta_0 = 0.001, beta_1 = 0.01) {
  frob_squared <- function(Fa, Fb) {
    sum((Fa - Fb)^2)
  }

  T <- length(F_list)
  E <- matrix(NA, nrow = T, ncol = 2)

  for (t in seq_len(T)) {
    Ft <- F_list[[t]]

    d0 <- frob_squared(Ft, F_0)
    d1 <- frob_squared(Ft, F_1)

    E[t, 1] <- exp(-beta_0 * d0)  # P(F_t | S_t = 0)
    E[t, 2] <- exp(-beta_1 * d1)  # P(F_t | S_t = 1)
  }

  # Safe normalization
  row_sums <- rowSums(E)
  row_sums[row_sums == 0] <- 1e-12  # avoid NaN
  E <- E / row_sums

  return(E)
}
```

## Apply on validation set

```{r}
# Assuming you already have:
#   - test_set (list of trees)
#   - F_0, F_1 (average F-matrices from training)
#   - beta_1 (start with 1.0)

# Prepare padded F-matrices from validation set
max_dim <- max(nrow(F_0), nrow(F_1))
pad_Fmat <- function(Fmat, target_dim) {
  padded <- matrix(0, nrow = target_dim, ncol = target_dim)
  padded[1:nrow(Fmat), 1:ncol(Fmat)] <- Fmat
  return(padded)
}

# F_val_list <- lapply(val_set, function(x) pad_Fmat(x$Fmat, max_dim))
F_val_list <- lapply(unlist(val_set, recursive = FALSE), function(x) pad_Fmat(x$Fmat, max_dim))

# Example: compute emission matrix with current beta values
E_val <- compute_emission_matrix(F_val_list, F_0, F_1, beta_0 = 0.001, beta_1 = 0.001)
```

## Visualize E_val statistics

```{r}
summary(E_val[,1])  # Neutral emission probs
summary(E_val[,2])  # Selective emission probs

# For all F-matrices in the validation set, What are the likelihoods of those matrices under each state?
hist(E_val[,1], breaks = 50, col = "#7cc4b0", main = "Emission probabilities for state = 0 (neutral)")
hist(E_val[,2], breaks = 50, col = "#7c9cc4", main = "Emission probabilities for state = 1 (selective)")
```

## Estimate Transition Matrix from training data

```{r}
# Initialize count matrix
trans_counts <- matrix(0, nrow = 2, ncol = 2)

for (chrom in train_set) {
  state_seq <- sapply(chrom, function(x) x$state)
  
  for (i in 1:(length(state_seq) - 1)) {
    from <- state_seq[i] + 1
    to   <- state_seq[i + 1] + 1
    trans_counts[from, to] <- trans_counts[from, to] + 1
  }
}

# Normalize rows to get transition probabilities
A_estimated <- trans_counts / rowSums(trans_counts)

# Result
print(A_estimated)
```

## Forward Algorithm

```{r}
forward_algorithm <- function(E, pi, A) {
  T <- nrow(E)
  alpha <- matrix(0, nrow = T, ncol = 2)

  alpha[1, ] <- pi * E[1, ]

  for (t in 2:T) {
    for (s in 1:2) {
      alpha[t, s] <- E[t, s] * sum(alpha[t - 1, ] * A[, s])
    }
  }

  L <- sum(alpha[T, ])
  if (L == 0) L <- 1e-300  # prevent log(0)
  return(list(alpha = alpha, log_likelihood = log(L)))
}
```

## Negative Log-Likelihood function for optim

```{r}
# ### Original ###
neg_log_likelihood <- function(beta_vec, F_list, F_0, F_1, pi, A) {
  beta_0 <- beta_vec[1]
  beta_1 <- beta_vec[2]

  # Compute emission matrix
  E <- compute_emission_matrix(F_list, F_0, F_1, beta_0, beta_1)

  # Run forward algorithm
  result <- forward_algorithm(E, pi, A)

  # Return negative log-likelihood
  return(-result$log_likelihood)
}
# ################

### With Regularization ###
# neg_log_likelihood <- function(beta_vec, F_list, F_0, F_1, pi, A, lambda = 1) {
#   beta_0 <- beta_vec[1]
#   beta_1 <- beta_vec[2]
# 
#   # Compute emission matrix
#   E <- compute_emission_matrix(F_list, F_0, F_1, beta_0, beta_1)
# 
#   # Run forward algorithm
#   result <- forward_algorithm(E, pi, A)
# 
#   # Regularization term
#   reg_term <- lambda * (beta_1 - beta_0)^2
# 
#   # Return regularized loss
#   return(-result$log_likelihood + reg_term)
# }
############################
```

## Apply algorithm

```{r}
# # Pad validation set F-matrices
# max_dim <- max(nrow(F_0), nrow(F_1))
# pad_Fmat <- function(Fmat, target_dim) {
#   padded <- matrix(0, nrow = target_dim, ncol = target_dim)
#   padded[1:nrow(Fmat), 1:ncol(Fmat)] <- Fmat
#   return(padded)
# }
# # pad_Fmat <- function(Fmat, target_dim) {
# #   m <- nrow(Fmat)
# #   if (m == target_dim) return(Fmat)
# #   
# #   padded <- matrix(0, nrow = target_dim, ncol = target_dim)  # define here
# #   padded[1:m, 1:m] <- Fmat
# #   return(padded)
# # }
# val_trees <- unlist(val_set, recursive = FALSE)
# F_val_list <- lapply(val_trees, function(x) pad_Fmat(x$Fmat, max_dim))
# F_val_list <- Filter(function(x) is.matrix(x), F_val_list)
# 
# # Initial guesses for beta values
# init_beta <- c(0.001, 0.002) # beta_0, beta_1
# 
# set.seed(1)
# F_val_subset <- sample(F_val_list, 5)
# 
# # Optimization
# opt_result <- optim(
#   par = init_beta,
#   fn = neg_log_likelihood,
#   method = "L-BFGS-B",
#   lower = c(1e-6, 1e-6),
#   upper = c(10, 10),
#   F_list = F_val_list,
#   F_0 = F_0,
#   F_1 = F_1,
#   pi = c(0.99, 0.01),
#   A = A_estimated,
#   # lambda = 1
# )
# 
# print(opt_result)

# Pad validation set F-matrices
max_dim <- max(nrow(F_0), nrow(F_1))

pad_Fmat <- function(Fmat, target_dim) {
  padded <- matrix(0, nrow = target_dim, ncol = target_dim)
  padded[1:nrow(Fmat), 1:ncol(Fmat)] <- Fmat
  return(padded)
}

val_trees <- unlist(val_set, recursive = FALSE)
F_val_list <- lapply(val_trees, function(x) pad_Fmat(x$Fmat, max_dim))
F_val_list <- Filter(function(x) is.matrix(x), F_val_list)


# Initial guesses for beta values
init_beta <- c(0.001, 0.001)  # beta_0, beta_1

# Optimization
opt_result <- optim(
  par = init_beta,
  fn = neg_log_likelihood,
  method = "L-BFGS-B",
  lower = c(1e-6, 1e-6),
  upper = c(10, 10),
  F_list = F_val_list,
  F_0 = F_0,
  F_1 = F_1,
  pi = c(0.99, 0.01),
  A = A_estimated,
  # lambda = 1
)

print(opt_result)
```

## Visualize optimized output

```{r}
beta_0 <- opt_result$par[1]
beta_1 <- opt_result$par[2]

E_val <- compute_emission_matrix(F_val_list, F_0, F_1, beta_0, beta_1)

hist(E_val[, 1], breaks = 50, col = "#7cc4b0", main = "P(F | Neutral)")
hist(E_val[, 2], breaks = 50, col = "#7c9cc4", main = "P(F | Selective)")
```

```{r}
log_ratio <- log(E_val[, 2] / E_val[, 1])
hist(log_ratio, breaks = 50, main = "Log-likelihood ratio: log(P(F|1)/P(F|0))")
```

## Test on Test set

```{r}
# beta_0 <- 0.002
# beta_1 <- 0.001

max_dim <- max(nrow(F_0), nrow(F_1))

pad_Fmat <- function(Fmat, target_dim) {
  padded <- matrix(0, nrow = target_dim, ncol = target_dim)
  padded[1:nrow(Fmat), 1:ncol(Fmat)] <- Fmat
  return(padded)
}
```

```{r}
posterior_list <- list()

for (i in seq_along(test_set)) {
  chrom <- test_set[[i]]
  
  F_list <- lapply(chrom, function(x) pad_Fmat(x$Fmat, max_dim))
  E <- compute_emission_matrix(F_list, F_0, F_1, beta_0, beta_1)

  # Run forward-backward to get posterior probabilities
  A <- A_estimated
  pi <- c(0.95, 0.05)

  forward_backward <- function(E, pi, A) {
    T <- nrow(E)
    N <- ncol(E)
    alpha <- matrix(0, T, N)
    beta <- matrix(0, T, N)
    
    # Forward pass
    alpha[1, ] <- pi * E[1, ]
    for (t in 2:T) {
      for (j in 1:N) {
        alpha[t, j] <- E[t, j] * sum(alpha[t - 1, ] * A[, j])
      }
    }
    
    # Backward pass
    beta[T, ] <- 1
    for (t in (T - 1):1) {
      for (i in 1:N) {
        beta[t, i] <- sum(A[i, ] * E[t + 1, ] * beta[t + 1, ])
      }
    }
    
    posterior <- alpha * beta
    posterior <- posterior / rowSums(posterior)
    return(posterior)
  }

  posterior_probs <- forward_backward(E, pi, A)
  probs_selective <- posterior_probs[, 2]

  df <- data.frame(
    pos = sapply(chrom, function(x) (x$left + x$right) / 2),
    prob = probs_selective,
    chromosome = i
  )

  posterior_list[[i]] <- df
}
```

```{r}
posterior_df <- do.call(rbind, posterior_list)

ggplot(posterior_df, aes(x = pos, y = prob, group = chromosome)) +
  geom_point(alpha = 0.5, color = "#7c9cc4", size = 0.8) +
  geom_vline(xintercept = 25000, color = "red") +
  labs(
    x = "Chromosome position",
    y = "Posterior P(Selective)",
    title = "Selection Probability Across Chromosomes"
  ) +
  theme_minimal()
```






