---
title: "load_and_store_data_new"
author: "Shinnosuke Yagi"
date: "2026-01-26"
output: html_document
---

## Overview

This script is a streamlined, memory-conscious variant of `load_and_store_data.Rmd`.
It only processes scenarios **r400, r404, r405, r406**, computes **F-matrices** and
**chromosome-wide weighted-average F-matrices**, and stores outputs in `data2/`.
No FW matrices are computed.

## Setup

```{r}
# # if phylodyn is attached
# if ("package:phylodyn" %in% search()) detach("package:phylodyn", unload = TRUE)
# 
# # also try unloading the namespace (in case it was imported but not attached)
# if ("phylodyn" %in% loadedNamespaces()) unloadNamespace("phylodyn")
# 
# remove.packages("phylodyn")
```

```{r, message=FALSE}
# One-time installs for phylodyn and dependencies
install.packages(c("ape", "spam", "devtools"))
# install.packages(
#   "INLA",
#   repos = c(getOption("repos"), INLA = "https://inla.r-inla-download.org/R/stable"),
#   dep = TRUE
# )

library(devtools)
install.packages("expm")
install_github("JuliaPalacios/phylodyn")
```

```{r, message=FALSE}
pkgs <- c(
  "ape",
  "spam",
  "readr",
  "dplyr",
  "stringr",
  "purrr",
  "phylodyn",
  "future.apply",
  "progressr",
  "future"
)
to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install) > 0) {
  install.packages(to_install)
}

library(ape)
library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(devtools)
# install_github("JuliaPalacios/phylodyn")
library(phylodyn)
library(future.apply)
library(progressr)
library(future)

source("load_and_store_data_new.R")

# --- Paths ---
# Directory containing *.tree and *_breaks.csv files
breaks_dir <- "../../Simulation/Slim(primary)/trees_newick"

# Output directory for results
out_dir <- "data2"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

# Parallel settings (adjust as needed)
use_parallel <- TRUE
# physical cores (better than logical for R multisession)
pc <- parallel::detectCores(logical = FALSE)
pc
n_workers <- max(1L, floor(min(pc - 1, 32) / 2))

# Configure future plan once for the session
if (use_parallel) {
  future::plan(future::multisession, workers = n_workers)
}
```

## Define scenarios

```{r}
# Scenario patterns (only these are processed)
patterns <- list(
  neutrals_r400       = "^r400.*\\.tree$",
  selectives_r404     = "^r404.*\\.tree$",
  selectives_r405     = "^r405.*\\.tree$",
  selectives_r406     = "^r406.*\\.tree$"
)
```

## Process scenarios

```{r}
# The function below processes one pattern at a time:
# - Reads each .tree file independently (no giant in-memory list)
# - Computes F-matrices for that file
# - Computes chromosome-wide weighted-average F
# - Saves both to data2/

results <- purrr::imap(patterns, function(pat, name) {
  cat("\n=== Processing:", name, "===\n")
  process_pattern_to_Fs(
    pattern = pat,
    data_dir = breaks_dir,
    out_dir = out_dir,
    scenario_name = name,
    use_parallel = use_parallel,
    n_workers = n_workers,
    show_tree_progress = FALSE,
    show_file_progress = TRUE
  )
})

# results is a named list of summaries for each scenario
results
```

## Empirical mean tree (neutral, first 250 weighted-average trees)

```{r}
# Compute the average F-matrix from the first 250 neutral weighted-average Fs
neutral_weighted_path <- file.path(out_dir, "F_weighted_average_neutrals_r400.rds")
neutral_weighted_list <- readRDS(neutral_weighted_path)

target_n <- 250L
count <- 0L
F_sum <- NULL

for (F in neutral_weighted_list) {
  if (is.null(F)) next
  if (is.null(F_sum)) {
    F_sum <- matrix(0, nrow = nrow(F), ncol = ncol(F))
  }
  F_sum <- F_sum + F
  count <- count + 1L
  if (count >= target_n) break
}

if (count == 0L) {
  stop("No neutral weighted-average F-matrices found in ", neutral_weighted_path)
}
if (count < target_n) {
  warning("Only ", count, " neutral weighted-average F-matrices found; averaging available matrices.")
}

F_empirical_mean <- F_sum / count
mean_path <- file.path(out_dir, "F_empirical_mean_neutrals_r400_first250.rds")
saveRDS(F_empirical_mean, mean_path)
cat("Saved empirical mean F-matrix ->", mean_path, "\n")
```

## Notes

- Each `.tree` file is processed in isolation to avoid huge memory usage.
- Per-file F-matrices are saved in `data2/Fs_nested_<scenario>/<file>.rds`.
- Weighted-average F matrices are stored as a single list per scenario:
  `data2/F_weighted_average_<scenario>.rds`.
