---
title: "chr2_20samples D1F (Post-Burn-in)"
author: "Codex"
date: "2026-02-16"
output: html_document
---

## Overview

This notebook computes per-tree **D1F** distances to the theoretical Kingman mean F-matrix for each population using the ARGweaver outputs in the new organized layout.

Pipeline:
1. Discover population/run folders under `raw_data/data/runs`.
2. Build a Kingman theoretical-mean proxy with `n=20`.
3. For each population, iterate over tree-sequence files (`*.tree`) and keep only post-burn-in MCMC samples (`sample > 90`).
4. Compute per-tree `D1F = sum(abs(F_tree - F_mean_kingman))`.
5. Save per-sample and per-population outputs under `raw_data/data/results`.

## Setup

```{r, message=FALSE}
library(ape)
library(phylodyn)
library(readr)
library(dplyr)
library(stringr)
library(tibble)
library(fmatrix)
library(progress)

options(stringsAsFactors = FALSE)
```

```{r}
project_dir <- normalizePath(".", winslash = "/", mustWork = TRUE)
data_root <- file.path(project_dir, "raw_data", "data")
runs_root <- file.path(data_root, "runs")
results_root <- file.path(data_root, "results")

stopifnot(dir.exists(runs_root))
if (!dir.exists(results_root)) dir.create(results_root, recursive = TRUE)

# Problem-specific settings
target_f_dim <- 39L
n_kingman <- target_f_dim + 1L  # fmatrix::kingman_m(n) returns (n-1)x(n-1)
burn_in_max <- 90L

run_dirs <- list.dirs(runs_root, recursive = FALSE, full.names = TRUE)
run_dirs <- run_dirs[grepl("^argweaver_", basename(run_dirs))]

run_table <- tibble(
  run_dir = run_dirs,
  run_name = basename(run_dirs),
  run_tag = sub("^argweaver_", "", basename(run_dirs)),
  pop = str_match(sub("^argweaver_", "", basename(run_dirs)), "_([A-Z]{3})20$")[, 2]
) %>%
  filter(!is.na(pop)) %>%
  arrange(pop, run_name)

if (nrow(run_table) == 0) {
  stop("No run directories matching pattern argweaver_*_<POP>20 were found in raw_data/data/runs.")
}

target_pops <- unique(run_table$pop)

cat("Discovered populations:", paste(target_pops, collapse = ", "), "\n")
cat("Using burn-in cutoff: sample <=", burn_in_max, "excluded\n")
```

## Kingman `n=20` theoretical-mean proxy F-matrix

```{r}
out_mean_rds <- file.path(results_root, sprintf("F_mean_kingman_n%d_proxy.rds", n_kingman))
F_mean_kingman <- fmatrix::kingman_m(n = n_kingman)

if (!all(dim(F_mean_kingman) == c(target_f_dim, target_f_dim))) {
  stop(sprintf(
    "Expected Kingman matrix to be %dx%d but got %s.",
    target_f_dim, target_f_dim, paste(dim(F_mean_kingman), collapse = "x")
  ))
}

saveRDS(F_mean_kingman, out_mean_rds)

cat("Saved Kingman mean proxy F-matrix:", out_mean_rds, "\n")
```

## Compute per-tree D1F distances (post-burn-in only)

```{r}
dist_d1 <- function(F1, F2) {
  sum(abs(F1 - F2))
}

conform_fmat <- function(Fm, target_dim = 39L) {
  d <- dim(Fm)
  if (any(d < target_dim)) {
    stop(sprintf(
      "F-matrix is smaller than %dx%d (%s).",
      target_dim, target_dim, paste(d, collapse = "x")
    ))
  }
  if (any(d > target_dim)) {
    Fm <- Fm[seq_len(target_dim), seq_len(target_dim), drop = FALSE]
  }
  Fm
}

process_run <- function(run_dir, run_tag, pop) {
  trees_dir <- file.path(run_dir, "trees")
  breaks_dir <- file.path(results_root, run_tag, "breakpoints")
  out_dir <- file.path(results_root, run_tag, "d1f_postburnin")
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  stopifnot(dir.exists(trees_dir))
  stopifnot(dir.exists(breaks_dir))

  tree_files <- list.files(trees_dir, pattern = "_sample[0-9]+\\.tree$", full.names = TRUE)
  if (length(tree_files) == 0) {
    warning(sprintf("[%s] No .tree files found in %s", pop, trees_dir))
    return(tibble())
  }

  run_manifest <- tibble(
    tree_path = tree_files,
    stem = tools::file_path_sans_ext(basename(tree_files)),
    sample = as.integer(str_match(basename(tree_files), "sample([0-9]+)\\.tree$")[, 2]),
    breaks_path = file.path(breaks_dir, paste0(tools::file_path_sans_ext(basename(tree_files)), "_breaks.csv"))
  ) %>%
    filter(!is.na(sample)) %>%
    filter(sample > burn_in_max) %>%
    arrange(sample)

  if (nrow(run_manifest) == 0) {
    warning(sprintf("[%s] No post-burn-in .tree files (sample > %d) in %s", pop, burn_in_max, trees_dir))
    return(tibble())
  }

  missing_breaks <- !file.exists(run_manifest$breaks_path)
  if (any(missing_breaks)) {
    warning(sprintf(
      "[%s] Missing breaks files for %d sample(s). They will be skipped.",
      pop, sum(missing_breaks)
    ))
    run_manifest <- run_manifest[!missing_breaks, , drop = FALSE]
  }

  if (nrow(run_manifest) == 0) {
    warning(sprintf("[%s] All post-burn-in samples were skipped due to missing breaks files.", pop))
    return(tibble())
  }

  cat("Population:", pop, "| Run:", run_tag, "\n")
  cat("Post-burn-in samples to process:", nrow(run_manifest), "\n")

  all_results <- vector("list", nrow(run_manifest))

  for (j in seq_len(nrow(run_manifest))) {
    tree_path <- run_manifest$tree_path[j]
    breaks_path <- run_manifest$breaks_path[j]
    sample <- run_manifest$sample[j]
    stem <- run_manifest$stem[j]

    trees <- ape::read.tree(tree_path)
    if (inherits(trees, "phylo")) {
      trees <- structure(list(trees), class = "multiPhylo")
    }

    breaks_df <- readr::read_csv(breaks_path, show_col_types = FALSE) %>%
      arrange(tree_index)

    n_trees <- length(trees)
    pb <- progress::progress_bar$new(
      format = sprintf("Computing D1F (%s sample %d) [:bar] :current/:total (:percent) eta=:eta", pop, sample),
      total = n_trees,
      clear = FALSE,
      width = 60
    )

    d1f_vals <- vapply(seq_len(n_trees), function(i) {
      pb$tick()
      Fm <- phylodyn::gen_Fmat(trees[[i]], tol = 8L)
      Fm <- conform_fmat(Fm, target_dim = target_f_dim)
      dist_d1(Fm, F_mean_kingman)
    }, numeric(1))

    n_use <- min(length(d1f_vals), nrow(breaks_df))
    if (length(d1f_vals) != nrow(breaks_df)) {
      warning(sprintf(
        "Length mismatch for %s (trees=%d, breaks=%d). Truncating to %d.",
        stem, length(d1f_vals), nrow(breaks_df), n_use
      ))
    }

    res <- tibble(
      population = pop,
      run_tag = run_tag,
      sample = sample,
      file = breaks_df$file[seq_len(n_use)],
      tree_index = breaks_df$tree_index[seq_len(n_use)],
      left = breaks_df$left[seq_len(n_use)],
      right = breaks_df$right[seq_len(n_use)],
      mid = breaks_df$mid[seq_len(n_use)],
      d1f_distance = d1f_vals[seq_len(n_use)]
    )

    out_sample_csv <- file.path(out_dir, sprintf("%s_d1f_per_tree_postburnin.csv", stem))
    readr::write_csv(res, out_sample_csv)
    cat("Saved:", out_sample_csv, "\n")

    all_results[[j]] <- res
  }

  dplyr::bind_rows(all_results)
}

all_d1f <- purrr::pmap_dfr(
  list(run_table$run_dir, run_table$run_tag, run_table$pop),
  process_run
)

cat("\nTotal post-burn-in rows:", nrow(all_d1f), "\n")
```

## Save per-population aggregated outputs

```{r}
if (nrow(all_d1f) > 0) {
  by_pop <- split(all_d1f, all_d1f$population)
  out_pop_dir <- file.path(results_root, "d1f_postburnin_n20_by_population")
  if (!dir.exists(out_pop_dir)) dir.create(out_pop_dir, recursive = TRUE)

  for (pop in names(by_pop)) {
    out_csv <- file.path(out_pop_dir, sprintf("chr2_20samples_%s_d1f_per_tree_postburnin.csv", pop))
    readr::write_csv(by_pop[[pop]], out_csv)
    cat("Saved aggregated file:", out_csv, "\n")
  }
} else {
  warning("No post-burn-in D1F rows were produced.")
}
```

## Quick summary

```{r}
if (nrow(all_d1f) > 0) {
  summary_tbl <- all_d1f %>%
    group_by(population, sample) %>%
    summarise(
      n_trees = n(),
      mean_d1f = mean(d1f_distance, na.rm = TRUE),
      median_d1f = median(d1f_distance, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(population, sample)

  print(summary_tbl, n = 200)
}
```
