---
title: "chr2_50samples D1FW"
author: "Codex"
date: "2026-02-25"
output: html_document
---

## Overview

This notebook computes per-tree **D1FW** distances for chr2, 50 samples per population.

Pipeline:
1. Load each population's Newick (`.tree`) and interval CSV (`_breaks.csv`).
2. Compute per-tree `FW = Fmat0 * Wmat0` using `phylodyn:::gen.tr.data2`.
3. Build a shared chr2 reference FW matrix by window-size weighted averaging across all populations.
4. Compute per-tree `D1FW = sum(abs(FW_tree - FW_reference))` and save one CSV per population.

## Setup

```{r, message=FALSE}
library(ape)
library(phylodyn)
library(readr)
library(dplyr)
library(progress)

options(stringsAsFactors = FALSE)
```

```{r}
base_dir <- normalizePath(".", winslash = "/", mustWork = TRUE)
trees_dir <- file.path(base_dir, "trees")
results_dir <- file.path(base_dir, "results")

CHR <- 2L
out_ref_rds <- file.path(results_dir, sprintf("chr%d_50samples_FW_weighted_reference_allpops.rds", CHR))

dist_d1 <- function(M1, M2) {
  sum(abs(M1 - M2))
}

jitter_internal_edges <- function(tr, eps = 1e-12) {
  tr2 <- tr
  if (is.null(tr2$edge.length)) return(tr2)

  n_tip <- ape::Ntip(tr2)
  idx_internal <- which(tr2$edge[, 2] > n_tip)
  if (length(idx_internal) == 0) return(tr2)

  ord <- order(tr2$edge[idx_internal, 2], tr2$edge[idx_internal, 1])
  offsets <- numeric(length(idx_internal))
  offsets[ord] <- seq_along(ord)

  tr2$edge.length[idx_internal] <- pmax(
    tr2$edge.length[idx_internal] + eps * offsets,
    .Machine$double.eps
  )
  tr2
}

binarize_tree <- function(tr) {
  tr2 <- tr
  if (!ape::is.binary(tr2)) {
    tr2 <- ape::multi2di(tr2, random = FALSE)
  }
  if (is.null(tr2$edge.length)) {
    tr2$edge.length <- rep(1, nrow(tr2$edge))
  }
  tr2$edge.length[!is.finite(tr2$edge.length) | tr2$edge.length <= 0] <- .Machine$double.eps
  tr2
}

jitter_all_edges <- function(tr, eps = 1e-8) {
  tr2 <- tr
  idx <- as.numeric(seq_along(tr2$edge.length))
  mult <- ((idx * 1103515245 + 12345) %% 997) + 1
  tr2$edge.length <- pmax(tr2$edge.length + eps * mult, .Machine$double.eps)
  tr2
}

compute_fw_safe <- function(
  tr,
  tol = 8L,
  tol_retry = c(6L, 4L, 2L, 0L, 10L, 12L, 14L),
  jitter_eps = c(1e-12, 1e-10, 1e-8, 1e-6, 1e-4)
) {
  try_once <- function(tree_obj, use_tol) {
    try(phylodyn:::gen.tr.data2(tree_obj, tol = use_tol), silent = TRUE)
  }

  tr0 <- binarize_tree(tr)
  tols <- unique(c(as.integer(tol), as.integer(tol_retry)))
  last_msg <- ""
  tie_msg <- FALSE

  for (tt in tols) {
    res <- try_once(tr0, tt)
    if (!inherits(res, "try-error")) {
      return(list(ok = TRUE, FW = res$Fmat0 * res$Wmat0, msg = ""))
    }
    last_msg <- as.character(res)
    tie_msg <- grepl(
      "more than one coalescent event at a given time",
      last_msg,
      ignore.case = TRUE
    )
    if (!tie_msg) {
      return(list(ok = FALSE, FW = NULL, msg = last_msg))
    }
  }

  tr_j <- tr0
  for (eps in jitter_eps) {
    tr_j <- jitter_internal_edges(tr_j, eps = eps)
    tr_j <- jitter_all_edges(tr_j, eps = eps)
    for (tt in tols) {
      res <- try_once(tr_j, tt)
      if (!inherits(res, "try-error")) {
        return(list(
          ok = TRUE,
          FW = res$Fmat0 * res$Wmat0,
          msg = sprintf("recovered_with_jitter_eps_%g_tol_%d", eps, tt)
        ))
      }
      last_msg <- as.character(res)
    }
  }

  list(ok = FALSE, FW = NULL, msg = last_msg)
}

chr2_pops <- list.files(trees_dir, pattern = sprintf("^chr%d_50samples_.*\\.tree$", CHR))
chr2_pops <- gsub(sprintf("^chr%d_50samples_", CHR), "", chr2_pops)
chr2_pops <- gsub("\\.tree$", "", chr2_pops)
chr2_pops <- sort(unique(chr2_pops))

cat("Chr2 populations discovered:", paste(chr2_pops, collapse = ", "), "\n")

# Manually overrride population list
chr2_pops <- c("CEU", "GBR", "TSI", "JPT", "KHV", "CHB")
```

## Build shared chr2 FW reference (window-size weighted across all populations)

```{r}
if (length(chr2_pops) == 0) {
  stop("No chr2 population .tree files found in trees_dir.")
}

FW_weighted_sum <- NULL
total_weight <- 0
used_summary <- list()

for (pop in chr2_pops) {
  input_trees <- file.path(trees_dir, sprintf("chr%d_50samples_%s.trees", CHR, pop))
  newick_path <- file.path(trees_dir, sprintf("chr%d_50samples_%s.tree", CHR, pop))
  breaks_path <- file.path(results_dir, sprintf("chr%d_50samples_%s_breaks.csv", CHR, pop))

  if (!file.exists(input_trees) || !file.exists(newick_path) || !file.exists(breaks_path)) {
    warning(sprintf("Skipping %s for reference build: missing one or more inputs.", pop))
    next
  }

  trees <- ape::read.tree(newick_path)
  if (inherits(trees, "phylo")) {
    trees <- structure(list(trees), class = "multiPhylo")
  }

  breaks_df <- readr::read_csv(breaks_path, show_col_types = FALSE) %>%
    dplyr::arrange(tree_index)

  n_use <- min(length(trees), nrow(breaks_df))
  if (n_use == 0) next

  widths <- breaks_df$right[seq_len(n_use)] - breaks_df$left[seq_len(n_use)]
  valid <- is.finite(widths) & (widths > 0)
  valid_idx <- which(valid)
  if (length(valid_idx) == 0) {
    warning(sprintf("Skipping %s for reference build: no positive finite window sizes.", pop))
    next
  }

  pb <- progress::progress_bar$new(
    format = sprintf("Building FW ref chr2 (%s) [:bar] :current/:total (:percent) eta=:eta", pop),
    total = length(valid_idx),
    clear = FALSE,
    width = 60
  )

  fail_ref <- 0L
  for (i in valid_idx) {
    pb$tick()
    fw_out <- compute_fw_safe(trees[[i]], tol = 8L)
    if (!isTRUE(fw_out$ok)) {
      fail_ref <- fail_ref + 1L
      next
    }
    FWm <- fw_out$FW
    if (is.null(FW_weighted_sum)) {
      FW_weighted_sum <- matrix(0, nrow = nrow(FWm), ncol = ncol(FWm))
    }
    FW_weighted_sum <- FW_weighted_sum + widths[i] * FWm
    total_weight <- total_weight + widths[i]
  }

  used_summary[[pop]] <- c(
    trees = length(trees),
    breaks = nrow(breaks_df),
    n_use = n_use,
    valid_windows = length(valid_idx),
    fw_failures = fail_ref,
    total_width = sum(widths[valid_idx])
  )
}

if (is.null(FW_weighted_sum) || total_weight <= 0) {
  stop("Failed to build FW reference: no usable trees/windows found.")
}

FW_ref <- FW_weighted_sum / total_weight
saveRDS(FW_ref, out_ref_rds)

cat("Saved shared chr2 FW reference:", out_ref_rds, "\n")
cat("Total reference weight:", total_weight, "\n")
print(used_summary)
```

## Compute per-tree FW matrices and D1FW distances (per population)

```{r}
if (!exists("CHR", inherits = FALSE)) {
  CHR <- 2L
}

if (!exists("base_dir", inherits = FALSE)) {
  base_dir <- normalizePath(".", winslash = "/", mustWork = TRUE)
}
if (!exists("trees_dir", inherits = FALSE)) {
  trees_dir <- file.path(base_dir, "trees")
}
if (!exists("results_dir", inherits = FALSE)) {
  results_dir <- file.path(base_dir, "results")
}
if (!exists("out_ref_rds", inherits = FALSE)) {
  out_ref_rds <- file.path(results_dir, sprintf("chr%d_50samples_FW_weighted_reference_allpops.rds", CHR))
}

if (!exists("dist_d1", mode = "function")) {
  dist_d1 <- function(M1, M2) sum(abs(M1 - M2))
}

if (!exists("compute_fw_safe", mode = "function")) {
  stop("compute_fw_safe() is not defined. Run the setup chunk that defines helper functions first.")
}

if (!exists("FW_ref", inherits = FALSE)) {
  if (!file.exists(out_ref_rds)) {
    stop(sprintf(
      "FW_ref is not in memory and reference RDS is missing: %s\nRun the reference-build chunk first or provide this file.",
      out_ref_rds
    ))
  }
  FW_ref <- readRDS(out_ref_rds)
}

if (!exists("chr2_pops", inherits = FALSE) || length(chr2_pops) == 0) {
  chr2_pops <- list.files(trees_dir, pattern = sprintf("^chr%d_50samples_.*\\.tree$", CHR))
  chr2_pops <- gsub(sprintf("^chr%d_50samples_", CHR), "", chr2_pops)
  chr2_pops <- gsub("\\.tree$", "", chr2_pops)
  chr2_pops <- sort(unique(chr2_pops))
}

process_pop_chr2_d1fw <- function(pop) {
  input_trees <- file.path(trees_dir, sprintf("chr%d_50samples_%s.trees", CHR, pop))
  newick_path <- file.path(trees_dir, sprintf("chr%d_50samples_%s.tree", CHR, pop))
  breaks_path <- file.path(results_dir, sprintf("chr%d_50samples_%s_breaks.csv", CHR, pop))
  out_csv <- file.path(results_dir, sprintf("chr%d_50samples_%s_d1fw_per_tree_weighted_ref.csv", CHR, pop))

  if (!file.exists(input_trees) || !file.exists(newick_path) || !file.exists(breaks_path)) {
    warning(sprintf("Skipping %s: missing one or more input files for chr2 D1FW.", pop))
    return(invisible(NULL))
  }

  trees <- ape::read.tree(newick_path)
  if (inherits(trees, "phylo")) {
    trees <- structure(list(trees), class = "multiPhylo")
  }

  breaks_df <- readr::read_csv(breaks_path, show_col_types = FALSE) %>%
    dplyr::arrange(tree_index)

  n_use <- min(length(trees), nrow(breaks_df))
  if (n_use == 0) {
    warning(sprintf("Skipping %s: no trees/break rows available.", pop))
    return(invisible(NULL))
  }

  if (length(trees) != nrow(breaks_df)) {
    warning(sprintf(
      "Length mismatch for chr2 D1FW %s (trees=%d, breaks=%d). Truncating to %d.",
      pop, length(trees), nrow(breaks_df), n_use
    ))
  }

  widths <- breaks_df$right[seq_len(n_use)] - breaks_df$left[seq_len(n_use)]
  valid <- is.finite(widths) & (widths > 0)

  pb <- progress::progress_bar$new(
    format = sprintf("Computing D1FW chr2 (%s) [:bar] :current/:total (:percent) eta=:eta", pop),
    total = n_use,
    clear = FALSE,
    width = 60
  )

  d1fw_vals <- rep(NA_real_, n_use)
  fw_failures <- 0L

  for (i in seq_len(n_use)) {
    pb$tick()
    fw_out <- compute_fw_safe(trees[[i]], tol = 8L)
    if (!isTRUE(fw_out$ok)) {
      fw_failures <- fw_failures + 1L
      next
    }
    d1fw_vals[i] <- dist_d1(fw_out$FW, FW_ref)
  }

  results <- tibble::tibble(
    file = breaks_df$file[seq_len(n_use)],
    tree_index = breaks_df$tree_index[seq_len(n_use)],
    left = breaks_df$left[seq_len(n_use)],
    right = breaks_df$right[seq_len(n_use)],
    mid = breaks_df$mid[seq_len(n_use)],
    window_size = widths,
    reference_type = "all_population_weighted_mean_FW",
    population = pop,
    d1fw_distance = d1fw_vals
  )

  results <- results %>%
    dplyr::filter(valid)

  readr::write_csv(results, out_csv)
  cat("Saved per-tree D1FW:", out_csv, "\n")
  cat("Population:", pop, "| Trees used:", n_use, "| FW failures:", fw_failures,
      "| Valid windows:", sum(valid), "\n\n")

  invisible(results)
}

d1fw_results <- lapply(chr2_pops, process_pop_chr2_d1fw)
```

