---
title: "chr2 signed D1F for 52-tip population trees"
author: "Codex"
date: "2026-03-16"
output: html_document
---

## Overview

This notebook computes the signed distance

\[
D1F_{signed}(x, T_k)
\]

for chr2 trees from:

- `CEU`
- `GBR`
- `TSI`
- `JPT`
- `KHV`
- `CHB`
- `allpop` (2 individuals from each of 26 populations; 52 tips total)

The sign rule is:

- negative if `D1F(x, T_unb) <= D1F(x, T_bal)`
- positive otherwise

The workflow reuses saved raw F-matrix caches in `results/fmatrix_cache` whenever they already exist, so raw per-tree F-matrices do not need to be recomputed unnecessarily.

## Setup

```{r, message=FALSE}
# Load packages used for tree reading, F-matrix generation, and CSV/RDS I/O.
library(ape)
library(phylodyn)
library(readr)
library(dplyr)
library(fmatrix)
library(progress)
library(tibble)

options(stringsAsFactors = FALSE)
```

## Paths and target definitions

```{r}
# Define the working directories and the chr2 target tree sets to process.
base_dir <- normalizePath(".", winslash = "/", mustWork = TRUE)
trees_dir <- file.path(base_dir, "trees")
results_dir <- file.path(base_dir, "results")
cache_dir <- file.path(results_dir, "fmatrix_cache")

if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}

CHR <- 2L
N_TIPS <- 52L
F_DIM_TARGET <- N_TIPS - 1L

target_specs <- tibble::tribble(
  ~target_label, ~source_stem,                 ~cache_stem,               ~output_stem,
  "CEU",         "chr2_52samples_CEU",         "chr2_52samples_CEU",      "chr2_52samples_CEU",
  "GBR",         "chr2_52samples_GBR",         "chr2_52samples_GBR",      "chr2_52samples_GBR",
  "TSI",         "chr2_52samples_TSI",         "chr2_52samples_TSI",      "chr2_52samples_TSI",
  "JPT",         "chr2_52samples_JPT",         "chr2_52samples_JPT",      "chr2_52samples_JPT",
  "KHV",         "chr2_52samples_KHV",         "chr2_52samples_KHV",      "chr2_52samples_KHV",
  "CHB",         "chr2_52samples_CHB",         "chr2_52samples_CHB",      "chr2_52samples_CHB",
  "allpop",      "chr2_2inds_allpops",         "chr2_allpops_2inds",      "chr2_2inds_allpops"
)

combined_out_csv <- file.path(results_dir, "chr2_signed_d1f_per_tree_selected_targets.csv")
reference_out_rds <- file.path(results_dir, "chr2_signed_d1f_reference_matrices_n52.rds")
manifest_out_csv <- file.path(results_dir, "chr2_signed_d1f_output_manifest.csv")

print(target_specs)
```

## Helper functions for references and distances

```{r}
# Compute the unsigned D1F distance between two F matrices.
dist_d1 <- function(F1, F2) {
  sum(abs(F1 - F2))
}

# Keep only the common 51 x 51 block used for all 52-tip comparisons.
trim_F_to_target <- function(Fm, target_dim = F_DIM_TARGET) {
  if (!is.matrix(Fm)) {
    stop("F matrix must be a matrix.")
  }
  if (nrow(Fm) < target_dim || ncol(Fm) < target_dim) {
    stop(sprintf(
      "F matrix is smaller than target (%d x %d): got %d x %d",
      target_dim, target_dim, nrow(Fm), ncol(Fm)
    ))
  }
  Fm[seq_len(target_dim), seq_len(target_dim), drop = FALSE]
}

# Convert supported objects from the fmatrix package into a plain matrix.
as_reference_F <- function(obj, tol = 8L) {
  if (is.matrix(obj)) {
    return(obj)
  }
  if (inherits(obj, "phylo")) {
    return(phylodyn::gen_Fmat(obj, tol = tol))
  }
  if (is.list(obj) && !is.null(obj$F) && is.matrix(obj$F)) {
    return(obj$F)
  }
  stop("Unsupported object returned by reference-tree function.")
}

# Call an fmatrix constructor with either named or positional n, depending on the function signature.
call_reference_constructor <- function(fun_name, n_tips) {
  fun <- get(fun_name, envir = asNamespace("fmatrix"))

  out <- tryCatch(
    fun(n = n_tips),
    error = function(e_named) {
      tryCatch(
        fun(n_tips),
        error = function(e_positional) {
          stop(sprintf(
            "Failed to call fmatrix::%s with n=%d using named or positional arguments.",
            fun_name, n_tips
          ))
        }
      )
    }
  )

  out
}

# Apply the sign rule based on proximity to the unbalanced vs balanced reference.
signed_d1f_to_kingman <- function(F_tree, F_kingman, F_unb, F_bal) {
  d_kingman <- dist_d1(F_tree, F_kingman)
  d_unb <- dist_d1(F_tree, F_unb)
  d_bal <- dist_d1(F_tree, F_bal)

  signed_val <- if (d_unb <= d_bal) -d_kingman else d_kingman

  list(
    d1f_signed = signed_val,
    d1f_kingman = d_kingman,
    d1f_unbalanced = d_unb,
    d1f_balanced = d_bal,
    closer_to = if (d_unb <= d_bal) "unbalanced" else "balanced"
  )
}
```

## Helper functions for tree inputs and cache reuse

```{r}
# Read a Newick file and return a multiPhylo object.
read_multi_phylo <- function(newick_path) {
  trees <- ape::read.tree(newick_path)
  if (inherits(trees, "phylo")) {
    trees <- structure(list(trees), class = "multiPhylo")
  }
  trees
}

# Load the accompanying interval table and keep rows in tree order.
read_breaks_df <- function(breaks_path) {
  readr::read_csv(breaks_path, show_col_types = FALSE) %>%
    dplyr::arrange(tree_index)
}

# Return cache path candidates; for allpop we support both cache names already used in the repo.
cache_candidates_for_target <- function(spec) {
  if (identical(spec$target_label, "allpop")) {
    return(c(
      file.path(cache_dir, "chr2_allpops_2inds_raw_fmats.rds"),
      file.path(cache_dir, "chr2_2inds_allpops_raw_fmats.rds")
    ))
  }

  file.path(cache_dir, sprintf("%s_raw_fmats.rds", spec$cache_stem))
}

# Check whether an existing cache contains usable matrices for the requested number of trees.
cache_valid <- function(obj, n_need, target_dim) {
  F_list <- NULL

  if (is.list(obj) && !is.null(obj$F_list) && is.list(obj$F_list)) {
    F_list <- obj$F_list
  } else if (is.list(obj)) {
    F_list <- obj
  }

  if (is.null(F_list) || length(F_list) < n_need) {
    return(FALSE)
  }

  idx <- seq_len(n_need)
  all(vapply(idx, function(i) {
    m <- F_list[[i]]
    is.matrix(m) && nrow(m) >= target_dim && ncol(m) >= target_dim
  }, logical(1)))
}

# Normalize either cache layout used in this repository into a single object with F_list + breaks.
normalize_cached_object <- function(cached, breaks_df, n_use, target_dim) {
  if (!cache_valid(cached, n_use, target_dim)) {
    stop("Cache is not valid for the requested number of trees.")
  }

  if (is.list(cached) && !is.null(cached$F_list) && is.list(cached$F_list)) {
    left <- cached$left
    right <- cached$right
    F_list <- cached$F_list
  } else {
    left <- breaks_df$left
    right <- breaks_df$right
    F_list <- cached
  }

  list(
    left = as.numeric(left[seq_len(n_use)]),
    right = as.numeric(right[seq_len(n_use)]),
    F_list = lapply(F_list[seq_len(n_use)], trim_F_to_target, target_dim = target_dim)
  )
}

# Reuse an existing raw cache when possible; otherwise compute it once and save it.
load_or_build_f_cache <- function(spec, force_recompute = FALSE, tol = 8L) {
  newick_path <- file.path(trees_dir, sprintf("%s.tree", spec$source_stem))
  breaks_path <- file.path(results_dir, sprintf("%s_breaks.csv", spec$source_stem))

  if (!file.exists(newick_path) || !file.exists(breaks_path)) {
    stop(sprintf(
      "Missing input(s) for %s: %s | %s",
      spec$target_label, newick_path, breaks_path
    ))
  }

  breaks_df <- read_breaks_df(breaks_path)
  trees <- NULL
  n_use <- NA_integer_

  if (!isTRUE(force_recompute)) {
    for (candidate in cache_candidates_for_target(spec)) {
      if (!file.exists(candidate)) {
        next
      }

      cached <- readRDS(candidate)

      if (is.null(trees)) {
        trees <- read_multi_phylo(newick_path)
        n_use <- min(length(trees), nrow(breaks_df))
      }

      if (n_use == 0) {
        stop(sprintf("No overlapping trees and break rows for %s.", spec$target_label))
      }

      if (cache_valid(cached, n_use, F_DIM_TARGET)) {
        obj <- normalize_cached_object(cached, breaks_df, n_use, F_DIM_TARGET)
        attr(obj, "cache_path_used") <- candidate
        attr(obj, "cache_source") <- "existing"
        return(obj)
      }
    }
  }

  if (is.null(trees)) {
    trees <- read_multi_phylo(newick_path)
    n_use <- min(length(trees), nrow(breaks_df))
  }

  if (n_use == 0) {
    stop(sprintf("No overlapping trees and break rows for %s.", spec$target_label))
  }

  if (length(trees) != nrow(breaks_df)) {
    warning(sprintf(
      "Length mismatch for %s (trees=%d, breaks=%d). Truncating to %d.",
      spec$target_label, length(trees), nrow(breaks_df), n_use
    ))
  }

  pb <- progress::progress_bar$new(
    format = sprintf("Building raw F cache (%s) [:bar] :current/:total (:percent) eta=:eta", spec$target_label),
    total = n_use,
    clear = FALSE,
    width = 60
  )

  F_list <- lapply(seq_len(n_use), function(i) {
    pb$tick()
    trim_F_to_target(phylodyn::gen_Fmat(trees[[i]], tol = tol), target_dim = F_DIM_TARGET)
  })

  obj <- list(
    left = as.numeric(breaks_df$left[seq_len(n_use)]),
    right = as.numeric(breaks_df$right[seq_len(n_use)]),
    F_list = F_list
  )

  save_path <- cache_candidates_for_target(spec)[1]
  saveRDS(obj, save_path)
  attr(obj, "cache_path_used") <- save_path
  attr(obj, "cache_source") <- "newly_built"
  obj
}
```

## Build the three 52-tip reference F-matrices

```{r}
# Construct and save the Kingman, balanced, and unbalanced reference F matrices for n = 52.
F_kingman <- trim_F_to_target(
  as_reference_F(call_reference_constructor("kingman_m", N_TIPS)),
  target_dim = F_DIM_TARGET
)

F_bal <- trim_F_to_target(
  as_reference_F(call_reference_constructor("bal_Fmat", N_TIPS)),
  target_dim = F_DIM_TARGET
)

F_unb <- trim_F_to_target(
  as_reference_F(call_reference_constructor("unb_Fmat", N_TIPS)),
  target_dim = F_DIM_TARGET
)

reference_bundle <- list(
  n_tips = N_TIPS,
  f_dim = F_DIM_TARGET,
  F_kingman = F_kingman,
  F_bal = F_bal,
  F_unb = F_unb
)

saveRDS(reference_bundle, reference_out_rds)
cat("Saved reference matrices:", reference_out_rds, "\n")
```

## Helper function to compute signed D1F for one target

```{r}
# Compute per-tree signed D1F for a single target and write the corresponding CSV.
compute_signed_d1f_for_target <- function(spec, force_recompute_cache = FALSE) {
  cache_obj <- load_or_build_f_cache(spec, force_recompute = force_recompute_cache)

  breaks_path <- file.path(results_dir, sprintf("%s_breaks.csv", spec$source_stem))
  breaks_df <- read_breaks_df(breaks_path)

  n_use <- min(length(cache_obj$F_list), nrow(breaks_df))
  if (n_use == 0) {
    stop(sprintf("No usable rows for %s.", spec$target_label))
  }

  pb <- progress::progress_bar$new(
    format = sprintf("Computing signed D1F (%s) [:bar] :current/:total (:percent) eta=:eta", spec$target_label),
    total = n_use,
    clear = FALSE,
    width = 60
  )

  score_df <- purrr::map_dfr(seq_len(n_use), function(i) {
    pb$tick()

    signed_obj <- signed_d1f_to_kingman(
      F_tree = cache_obj$F_list[[i]],
      F_kingman = F_kingman,
      F_unb = F_unb,
      F_bal = F_bal
    )

    tibble::tibble(
      chromosome = CHR,
      target = spec$target_label,
      source_stem = spec$source_stem,
      file = breaks_df$file[i],
      tree_index = breaks_df$tree_index[i],
      left = as.numeric(breaks_df$left[i]),
      right = as.numeric(breaks_df$right[i]),
      mid = as.numeric(breaks_df$mid[i]),
      d1f_signed = signed_obj$d1f_signed,
      d1f_kingman = signed_obj$d1f_kingman,
      d1f_unbalanced = signed_obj$d1f_unbalanced,
      d1f_balanced = signed_obj$d1f_balanced,
      sign_basis = signed_obj$closer_to,
      n_tips = N_TIPS,
      f_dim = F_DIM_TARGET,
      raw_f_cache = attr(cache_obj, "cache_path_used"),
      raw_f_cache_source = attr(cache_obj, "cache_source")
    )
  })

  out_csv <- file.path(results_dir, sprintf("%s_d1f_signed_per_tree.csv", spec$output_stem))
  readr::write_csv(score_df, out_csv)

  cat("Saved signed D1F output:", out_csv, "\n")
  cat("Raw F cache used:", attr(cache_obj, "cache_path_used"), "\n")

  score_df
}
```

## Compute signed D1F for the selected populations and allpop

```{r}
# Run the signed-D1F calculation for each requested target and save both per-target and combined outputs.
force_recompute_cache <- FALSE

all_results <- vector("list", nrow(target_specs))
output_manifest <- vector("list", nrow(target_specs))

for (i in seq_len(nrow(target_specs))) {
  spec <- target_specs[i, ]
  out_df <- compute_signed_d1f_for_target(spec, force_recompute_cache = force_recompute_cache)
  all_results[[i]] <- out_df

  output_manifest[[i]] <- tibble::tibble(
    target = spec$target_label,
    source_stem = spec$source_stem,
    output_csv = file.path(results_dir, sprintf("%s_d1f_signed_per_tree.csv", spec$output_stem))
  )
}

combined_df <- dplyr::bind_rows(all_results)
manifest_df <- dplyr::bind_rows(output_manifest)

readr::write_csv(combined_df, combined_out_csv)
readr::write_csv(manifest_df, manifest_out_csv)

cat("Saved combined signed D1F CSV:", combined_out_csv, "\n")
cat("Saved output manifest:", manifest_out_csv, "\n")

print(dplyr::slice_head(combined_df, n = 10))
```
