---
title: "chr2_50samples D1F"
author: "Codex"
date: "2026-02-12"
output: html_document
---

## Overview

This notebook computes per-tree **D1F** distances for multiple populations:
**CER, GBR, TSI, JPT**.

Pipeline:
1. Use pre-generated Newick (`.tree`) plus interval CSV (`_breaks.csv`) from `1000Genomes/load_1kg_tsz.ipynb`.
2. Load each tree and compute its F-matrix with `phylodyn::gen_Fmat`.
3. Build a Kingman `n=50` theoretical-mean proxy tree using expected coalescent waiting times and compute its F-matrix.
4. Compute per-tree `D1F = sum(abs(F_tree - F_mean_kingman))` and save results per population.

## Setup

```{r, message=FALSE}
library(ape)
library(phylodyn)
library(readr)
library(dplyr)
library(purrr)
library(fmatrix)
library(progress)

options(stringsAsFactors = FALSE)
```

```{r}
base_dir <- normalizePath(".", winslash = "/", mustWork = TRUE)

trees_dir <- file.path(base_dir, "trees")
results_dir <- file.path(base_dir, "results")

target_pops <- c("CEU", "GBR", "TSI", "JPT", "KHV", "CHB")

out_mean_rds <- file.path(results_dir, "F_mean_kingman_n50_proxy.rds")
```

## Kingman `n=50` theoretical-mean proxy tree and F-matrix

```{r}
F_mean_kingman <- fmatrix::kingman_m(n = 50L)
saveRDS(F_mean_kingman, out_mean_rds)

cat("Saved Kingman mean proxy F-matrix:", out_mean_rds, "
")
```

## Compute per-tree F-matrices and D1F distances (per population)

```{r}
dist_d1 <- function(F1, F2) {
  sum(abs(F1 - F2))
}

process_pop <- function(pop) {
  input_trees <- file.path(trees_dir, sprintf("chr2_50samples_%s.trees", pop))
  newick_path <- file.path(trees_dir, sprintf("chr2_50samples_%s.tree", pop))
  breaks_path <- file.path(results_dir, sprintf("chr2_50samples_%s_breaks.csv", pop))
  out_csv <- file.path(results_dir, sprintf("chr2_50samples_%s_d1f_per_tree.csv", pop))

  stopifnot(file.exists(input_trees))
  stopifnot(file.exists(newick_path))
  stopifnot(file.exists(breaks_path))

  trees <- ape::read.tree(newick_path)
  if (inherits(trees, "phylo")) {
    trees <- structure(list(trees), class = "multiPhylo")
  }

  breaks_df <- readr::read_csv(breaks_path, show_col_types = FALSE) %>%
    arrange(tree_index)

  cat("Population:", pop, "
")
  cat("Trees loaded:", length(trees), "
")
  cat("Break rows   :", nrow(breaks_df), "
")

  n_trees <- length(trees)

  pb <- progress::progress_bar$new(
    format = sprintf("Computing D1F (%s) [:bar] :current/:total (:percent) eta=:eta", pop),
    total = n_trees,
    clear = FALSE,
    width = 60
  )

  d1f_vals <- vapply(seq_len(n_trees), function(i) {
    pb$tick()
    Fm <- phylodyn::gen_Fmat(trees[[i]], tol = 8L)
    dist_d1(Fm, F_mean_kingman)
  }, numeric(1))

  n_use <- min(length(d1f_vals), nrow(breaks_df))
  if (length(d1f_vals) != nrow(breaks_df)) {
    warning(sprintf(
      "Length mismatch for %s (trees=%d, breaks=%d). Truncating to %d.",
      pop, length(d1f_vals), nrow(breaks_df), n_use
    ))
  }

  results <- tibble::tibble(
    file = breaks_df$file[seq_len(n_use)],
    tree_index = breaks_df$tree_index[seq_len(n_use)],
    left = breaks_df$left[seq_len(n_use)],
    right = breaks_df$right[seq_len(n_use)],
    mid = breaks_df$mid[seq_len(n_use)],
    d1f_distance = d1f_vals[seq_len(n_use)]
  )

  readr::write_csv(results, out_csv)
  cat("Saved per-tree D1F:", out_csv, "

")

  invisible(results)
}

results_list <- lapply(target_pops, process_pop)
```

