---
title: "chr2 52samples vs allpop(2inds) binned D1F"
author: "Codex"
date: "2026-03-03"
output: html_document
---

## Overview

This notebook computes, for each population and genomic bin, the D1F distance between:

1. A population-specific window-averaged F matrix from `chr2_<num_samples>samples_<POP>` trees.
2. An all-population window-averaged F matrix from `chr2_2inds_allpops` trees.

Default bin size is 5,000 bp.

## Setup

```{r, message=FALSE}
library(ape)
library(phylodyn)
library(readr)
library(dplyr)
library(progress)

options(stringsAsFactors = FALSE)
```

```{r}
base_dir <- normalizePath(".", winslash = "/", mustWork = TRUE)
trees_dir <- file.path(base_dir, "trees")
results_dir <- file.path(base_dir, "results")

CHR <- 2L
num_samples <- 52L
bin_size_bp <- 5000L

target_pops <- c("CEU", "GBR", "TSI", "JPT", "KHV", "CHB")
pop_tag <- sprintf("%dsamples", as.integer(num_samples))

allpops_stem <- sprintf("chr%d_2inds_allpops", CHR)
allpops_label <- "allpops_2inds"

cache_dir <- file.path(results_dir, "fmatrix_cache")
if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}

allpops_cache_rds <- file.path(
  cache_dir,
  sprintf("chr%d_%s_raw_fmats.rds", CHR, allpops_label)
)

out_csv <- file.path(
  results_dir,
  sprintf("chr%d_%s_vs_allpops2inds_d1f_between_pop_binned_bin%d.csv", CHR, pop_tag, as.integer(bin_size_bp))
)

cat("CHR:", CHR, "| pop_tag:", pop_tag, "| bin_size_bp:", bin_size_bp, "\n")
cat("Output CSV:", out_csv, "\n")
cat("Cache dir:", cache_dir, "\n")
```

## Helpers

```{r}
dist_d1 <- function(F1, F2) {
  sum(abs(F1 - F2))
}

trim_for_target_dim <- function(Fm, target_dim) {
  if (!is.matrix(Fm)) {
    stop("Expected F matrix.")
  }

  nr <- nrow(Fm)
  nc <- ncol(Fm)
  if (nr > target_dim || nc > target_dim) {
    if (nr < target_dim || nc < target_dim) {
      stop(sprintf(
        "Cannot trim matrix to %d x %d because current size is %d x %d.",
        target_dim, target_dim, nr, nc
      ))
    }
    return(Fm[seq_len(target_dim), seq_len(target_dim), drop = FALSE])
  }

  Fm
}

load_trees_and_breaks <- function(newick_path, breaks_path, label) {
  if (!file.exists(newick_path) || !file.exists(breaks_path)) {
    stop(sprintf("Missing input(s) for %s: %s | %s", label, newick_path, breaks_path))
  }

  trees <- ape::read.tree(newick_path)
  if (inherits(trees, "phylo")) {
    trees <- structure(list(trees), class = "multiPhylo")
  }

  breaks_df <- readr::read_csv(breaks_path, show_col_types = FALSE) %>%
    dplyr::arrange(tree_index)

  required <- c("left", "right", "mid")
  miss <- setdiff(required, names(breaks_df))
  if (length(miss) > 0) {
    stop(sprintf("%s missing columns in breaks file: %s", label, paste(miss, collapse = ", ")))
  }

  n_use <- min(length(trees), nrow(breaks_df))
  if (n_use == 0) {
    stop(sprintf("No usable trees/break rows for %s", label))
  }

  if (length(trees) != nrow(breaks_df)) {
    warning(sprintf(
      "Length mismatch for %s (trees=%d, breaks=%d). Truncating to %d.",
      label, length(trees), nrow(breaks_df), n_use
    ))
  }

  left <- as.numeric(breaks_df$left[seq_len(n_use)])
  right <- as.numeric(breaks_df$right[seq_len(n_use)])

  pb <- progress::progress_bar$new(
    format = sprintf("Computing F matrices (%s) [:bar] :current/:total (:percent) eta=:eta", label),
    total = n_use,
    clear = FALSE,
    width = 60
  )

  F_list <- vector("list", n_use)
  for (i in seq_len(n_use)) {
    pb$tick()
    F_list[[i]] <- phylodyn::gen_Fmat(trees[[i]], tol = 8L)
  }

  list(
    left = left,
    right = right,
    F_list = F_list
  )
}

binned_weighted_F <- function(obj, L, R) {
  overlap <- pmin(obj$right, R) - pmax(obj$left, L)
  keep <- which(is.finite(overlap) & overlap > 0)

  if (length(keep) == 0) {
    return(list(F = NULL, weight = 0, n_segments = 0L))
  }

  w <- overlap[keep]
  F0 <- obj$F_list[[keep[1]]]
  F_sum <- matrix(0, nrow = nrow(F0), ncol = ncol(F0))

  for (j in seq_along(keep)) {
    idx <- keep[j]
    F_sum <- F_sum + w[j] * obj$F_list[[idx]]
  }

  list(
    F = F_sum / sum(w),
    weight = sum(w),
    n_segments = as.integer(length(keep))
  )
}

binned_weighted_F_details <- function(obj, L, R, target_dim) {
  overlap <- pmin(obj$right, R) - pmax(obj$left, L)
  keep <- which(is.finite(overlap) & overlap > 0)

  if (length(keep) == 0) {
    return(list(
      F = NULL,
      weight = 0,
      n_segments = 0L,
      keep = integer(0),
      w = numeric(0)
    ))
  }

  w <- as.numeric(overlap[keep])
  F0 <- trim_for_target_dim(obj$F_list[[keep[1]]], target_dim)
  F_sum <- matrix(0, nrow = nrow(F0), ncol = ncol(F0))

  for (j in seq_along(keep)) {
    idx <- keep[j]
    Fm <- trim_for_target_dim(obj$F_list[[idx]], target_dim)
    if (nrow(Fm) != nrow(F0) || ncol(Fm) != ncol(F0)) {
      stop(sprintf(
        "F dimension mismatch inside a bin: base=%dx%d vs current=%dx%d",
        nrow(F0), ncol(F0), nrow(Fm), ncol(Fm)
      ))
    }
    F_sum <- F_sum + w[j] * Fm
  }

  list(
    F = F_sum / sum(w),
    weight = sum(w),
    n_segments = as.integer(length(keep)),
    keep = keep,
    w = w
  )
}

weighted_mean_distance_to_reference <- function(obj, details, target_dim, F_ref = NULL) {
  ref_use <- F_ref
  if (is.null(ref_use)) {
    ref_use <- details$F
  }

  if (is.null(ref_use) || length(details$keep) == 0) {
    return(NA_real_)
  }

  dvals <- vapply(seq_along(details$keep), function(j) {
    idx <- details$keep[j]
    Fm <- trim_for_target_dim(obj$F_list[[idx]], target_dim)
    if (nrow(Fm) != nrow(ref_use) || ncol(Fm) != ncol(ref_use)) {
      return(NA_real_)
    }
    dist_d1(Fm, ref_use)
  }, numeric(1))

  ok <- is.finite(dvals) & is.finite(details$w) & (details$w > 0)
  if (!any(ok)) {
    return(NA_real_)
  }
  stats::weighted.mean(dvals[ok], w = details$w[ok])
}
```

## Step 1: Build/cache raw F-matrices for all-population trees

```{r}
force_recompute_allpops_f <- FALSE

allpops_newick <- file.path(trees_dir, sprintf("%s.tree", allpops_stem))
allpops_breaks <- file.path(results_dir, sprintf("%s_breaks.csv", allpops_stem))

pb_step1 <- progress::progress_bar$new(
  format = "Step 1 (allpops cache) [:bar] :current/:total (:percent) eta=:eta",
  total = 1,
  clear = FALSE,
  width = 60
)

if (!file.exists(allpops_cache_rds) || isTRUE(force_recompute_allpops_f)) {
  allpops_obj <- load_trees_and_breaks(
    allpops_newick,
    allpops_breaks,
    label = allpops_label
  )
  saveRDS(allpops_obj, allpops_cache_rds)
  cat("Saved allpops raw F-matrix cache:", allpops_cache_rds, "\n")
} else {
  cat("Using existing allpops raw F-matrix cache:", allpops_cache_rds, "\n")
}
pb_step1$tick()
```

## Step 2: Build/cache raw F-matrices for each population tree set

```{r}
force_recompute_pop_f <- FALSE

pb_step2 <- progress::progress_bar$new(
  format = "Step 2 (population caches) [:bar] :current/:total (:percent) eta=:eta",
  total = length(target_pops),
  clear = FALSE,
  width = 60
)

for (pop in target_pops) {
  pop_cache_rds <- file.path(
    cache_dir,
    sprintf("chr%d_%s_%s_raw_fmats.rds", CHR, pop_tag, pop)
  )

  if (file.exists(pop_cache_rds) && !isTRUE(force_recompute_pop_f)) {
    cat("Using existing population raw F-matrix cache:", pop_cache_rds, "\n")
    pb_step2$tick()
    next
  }

  pop_newick <- file.path(trees_dir, sprintf("chr%d_%s_%s.tree", CHR, pop_tag, pop))
  pop_breaks <- file.path(results_dir, sprintf("chr%d_%s_%s_breaks.csv", CHR, pop_tag, pop))

  pop_obj <- load_trees_and_breaks(
    pop_newick,
    pop_breaks,
    label = sprintf("%s_%s", pop_tag, pop)
  )
  saveRDS(pop_obj, pop_cache_rds)
  cat("Saved population raw F-matrix cache:", pop_cache_rds, "\n")
  pb_step2$tick()
}
```

## Step 3: Compute per-bin D1F(pop vs allpops) using cached F-matrices

```{r}
if (!file.exists(allpops_cache_rds)) {
  stop(sprintf("Missing allpops cache: %s. Run Step 1 first.", allpops_cache_rds))
}
allpops_obj <- readRDS(allpops_cache_rds)
F_DIM_TARGET <- as.integer(num_samples) - 1L

results_all <- list()
pb_step3 <- progress::progress_bar$new(
  format = "Step 3 (binned pop-vs-allpops) [:bar] :current/:total (:percent) eta=:eta",
  total = length(target_pops),
  clear = FALSE,
  width = 60
)
for (pop in target_pops) {
  pop_cache_rds <- file.path(
    cache_dir,
    sprintf("chr%d_%s_%s_raw_fmats.rds", CHR, pop_tag, pop)
  )
  if (!file.exists(pop_cache_rds)) {
    stop(sprintf("Missing population cache for %s: %s. Run Step 2 first.", pop, pop_cache_rds))
  }
  pop_obj <- readRDS(pop_cache_rds)

  x0 <- min(min(pop_obj$left, na.rm = TRUE), min(allpops_obj$left, na.rm = TRUE))
  x1 <- max(max(pop_obj$right, na.rm = TRUE), max(allpops_obj$right, na.rm = TRUE))
  x0 <- floor(x0 / bin_size_bp) * bin_size_bp
  x1 <- ceiling(x1 / bin_size_bp) * bin_size_bp

  bin_edges <- seq(x0, x1, by = bin_size_bp)
  if (tail(bin_edges, 1) < x1) {
    bin_edges <- c(bin_edges, x1)
  }

  bin_left <- bin_edges[-length(bin_edges)]
  bin_right <- bin_edges[-1]
  bin_mid <- (bin_left + bin_right) / 2

  pb_bins <- progress::progress_bar$new(
    format = sprintf("Computing pop-vs-allpops D1F (%s) [:bar] :current/:total (:percent) eta=:eta", pop),
    total = length(bin_left),
    clear = FALSE,
    width = 60
  )

  pop_bin_rows <- vector("list", length(bin_left))
  for (i in seq_along(bin_left)) {
    pb_bins$tick()

    L <- bin_left[i]
    R <- bin_right[i]

    pop_ref <- binned_weighted_F(pop_obj, L, R)
    all_ref <- binned_weighted_F(allpops_obj, L, R)

    d_pop_all <- NA_real_
    if (!is.null(pop_ref$F) && !is.null(all_ref$F)) {
      F_pop_cmp <- trim_for_target_dim(pop_ref$F, F_DIM_TARGET)
      F_all_cmp <- trim_for_target_dim(all_ref$F, F_DIM_TARGET)

      if (nrow(F_pop_cmp) == nrow(F_all_cmp) && ncol(F_pop_cmp) == ncol(F_all_cmp)) {
        d_pop_all <- dist_d1(F_pop_cmp, F_all_cmp)
      } else {
        warning(sprintf(
          "Skipping bin [%g, %g) for %s due to F dimension mismatch after trimming (%dx%d vs %dx%d).",
          L, R, pop, nrow(F_pop_cmp), ncol(F_pop_cmp), nrow(F_all_cmp), ncol(F_all_cmp)
        ))
      }
    }

    pop_bin_rows[[i]] <- tibble::tibble(
      chromosome = CHR,
      population = pop,
      pop_sample_tag = pop_tag,
      baseline = allpops_label,
      bin_size_bp = as.integer(bin_size_bp),
      bin_left = as.numeric(L),
      bin_right = as.numeric(R),
      bin_mid = as.numeric(bin_mid[i]),
      n_overlap_pop = as.integer(pop_ref$n_segments),
      n_overlap_allpops = as.integer(all_ref$n_segments),
      overlap_bp_pop = as.numeric(pop_ref$weight),
      overlap_bp_allpops = as.numeric(all_ref$weight),
      d1f_distance_pop_vs_allpops = d_pop_all
    )
  }

  results_all[[pop]] <- dplyr::bind_rows(pop_bin_rows)
  pb_step3$tick()
}

out_df <- dplyr::bind_rows(results_all)
readr::write_csv(out_df, out_csv)

cat("Saved binned pop-vs-allpops D1F:", out_csv, "\n")
print(dplyr::slice_head(out_df, n = 10))
```

## Step 4: Compute local F_ST^Tree for CEU vs allpops (CEU/JPT-matched workflow)

```{r}
fst_pop <- "KHV"
out_fst_local_csv <- file.path(
  results_dir,
  sprintf("chr%d_%s_%s_vs_allpops2inds_fst_tree_local_bin%d.csv", CHR, pop_tag, fst_pop, as.integer(bin_size_bp))
)

if (!file.exists(allpops_cache_rds)) {
  stop(sprintf("Missing allpops cache: %s. Run Step 1 first.", allpops_cache_rds))
}
allpops_obj <- readRDS(allpops_cache_rds)

ceu_cache_rds <- file.path(
  cache_dir,
  sprintf("chr%d_%s_%s_raw_fmats.rds", CHR, pop_tag, fst_pop)
)
if (!file.exists(ceu_cache_rds)) {
  stop(sprintf("Missing CEU cache: %s. Run Step 2 first.", ceu_cache_rds))
}
ceu_obj <- readRDS(ceu_cache_rds)

F_DIM_TARGET <- as.integer(num_samples) - 1L

segment_width <- function(obj) {
  w <- as.numeric(obj$right - obj$left)
  w[!is.finite(w) | (w <= 0)] <- NA_real_
  w
}

build_weighted_reference <- function(obj, target_dim) {
  w <- segment_width(obj)
  keep <- which(is.finite(w) & (w > 0))
  if (length(keep) == 0) {
    stop("No positive-width segments for weighted reference.")
  }

  F0 <- trim_for_target_dim(obj$F_list[[keep[1]]], target_dim)
  F_sum <- matrix(0, nrow = nrow(F0), ncol = ncol(F0))
  w_sum <- 0
  for (idx in keep) {
    Fm <- trim_for_target_dim(obj$F_list[[idx]], target_dim)
    if (nrow(Fm) != nrow(F0) || ncol(Fm) != ncol(F0)) {
      stop(sprintf(
        "F dimension mismatch while building weighted reference: base=%dx%d vs current=%dx%d",
        nrow(F0), ncol(F0), nrow(Fm), ncol(Fm)
      ))
    }
    F_sum <- F_sum + w[idx] * Fm
    w_sum <- w_sum + w[idx]
  }
  F_sum / w_sum
}

distance_track_to_reference <- function(obj, F_ref, target_dim) {
  vapply(seq_along(obj$F_list), function(i) {
    Fm <- trim_for_target_dim(obj$F_list[[i]], target_dim)
    if (nrow(Fm) != nrow(F_ref) || ncol(Fm) != ncol(F_ref)) {
      return(NA_real_)
    }
    dist_d1(Fm, F_ref)
  }, numeric(1))
}

binned_weighted_mean_segments <- function(seg_left, seg_right, seg_val, bin_left, bin_right) {
  out_mean <- rep(NA_real_, length(bin_left))
  out_weight <- rep(0, length(bin_left))
  out_n <- rep(0L, length(bin_left))

  for (i in seq_along(bin_left)) {
    L <- bin_left[i]
    R <- bin_right[i]
    overlap <- pmin(seg_right, R) - pmax(seg_left, L)
    keep <- which(is.finite(overlap) & overlap > 0)
    out_n[i] <- as.integer(length(keep))
    if (length(keep) == 0) {
      next
    }

    w <- as.numeric(overlap[keep])
    out_weight[i] <- sum(w)

    v <- as.numeric(seg_val[keep])
    ok <- is.finite(v) & is.finite(w) & (w > 0)
    if (any(ok)) {
      out_mean[i] <- stats::weighted.mean(v[ok], w = w[ok])
    }
  }

  list(mean = out_mean, weight = out_weight, n_segments = out_n)
}

# CEU/JPT-style references:
# - within: per-tree distance to each group's chromosome-wide weighted reference
# - total : per-tree distance to grouped reference (mean of two weighted references)
F_ref_ceu <- build_weighted_reference(ceu_obj, F_DIM_TARGET)
F_ref_all <- build_weighted_reference(allpops_obj, F_DIM_TARGET)
if (nrow(F_ref_ceu) != nrow(F_ref_all) || ncol(F_ref_ceu) != ncol(F_ref_all)) {
  stop(sprintf(
    "Reference dimension mismatch (CEU=%dx%d vs allpops=%dx%d)",
    nrow(F_ref_ceu), ncol(F_ref_ceu), nrow(F_ref_all), ncol(F_ref_all)
  ))
}
F_ref_grouped <- (F_ref_ceu + F_ref_all) / 2

d_within_ceu <- distance_track_to_reference(ceu_obj, F_ref_ceu, F_DIM_TARGET)
d_within_all <- distance_track_to_reference(allpops_obj, F_ref_all, F_DIM_TARGET)
d_total_ceu <- distance_track_to_reference(ceu_obj, F_ref_grouped, F_DIM_TARGET)
d_total_all <- distance_track_to_reference(allpops_obj, F_ref_grouped, F_DIM_TARGET)

x0 <- min(min(ceu_obj$left, na.rm = TRUE), min(allpops_obj$left, na.rm = TRUE))
x1 <- max(max(ceu_obj$right, na.rm = TRUE), max(allpops_obj$right, na.rm = TRUE))
x0 <- floor(x0 / bin_size_bp) * bin_size_bp
x1 <- ceiling(x1 / bin_size_bp) * bin_size_bp

bin_edges <- seq(x0, x1, by = bin_size_bp)
if (tail(bin_edges, 1) < x1) {
  bin_edges <- c(bin_edges, x1)
}

bin_left <- bin_edges[-length(bin_edges)]
bin_right <- bin_edges[-1]
bin_mid <- (bin_left + bin_right) / 2

pb_step4 <- progress::progress_bar$new(
  format = "Step 4 (local F_ST^Tree CEU vs allpops) [:bar] :current/:total (:percent) eta=:eta",
  total = 1,
  clear = FALSE,
  width = 60
)

ceu_within_bin <- binned_weighted_mean_segments(
  ceu_obj$left, ceu_obj$right, d_within_ceu, bin_left, bin_right
)
all_within_bin <- binned_weighted_mean_segments(
  allpops_obj$left, allpops_obj$right, d_within_all, bin_left, bin_right
)
ceu_total_bin <- binned_weighted_mean_segments(
  ceu_obj$left, ceu_obj$right, d_total_ceu, bin_left, bin_right
)
all_total_bin <- binned_weighted_mean_segments(
  allpops_obj$left, allpops_obj$right, d_total_all, bin_left, bin_right
)
pb_step4$tick()

v_within <- 0.5 * (ceu_within_bin$mean + all_within_bin$mean)
v_total <- 0.5 * (ceu_total_bin$mean + all_total_bin$mean)
fst_raw <- 1 - (v_within / v_total)
fst_raw[!is.finite(v_total) | (v_total <= 0)] <- NA_real_
fst_clipped <- pmax(0, fst_raw)
fst_clipped[!is.finite(fst_raw)] <- NA_real_

fst_df <- tibble::tibble(
  chromosome = CHR,
  population = fst_pop,
  baseline = allpops_label,
  pop_sample_tag = pop_tag,
  bin_size_bp = as.integer(bin_size_bp),
  bin_left = as.numeric(bin_left),
  bin_right = as.numeric(bin_right),
  bin_mid = as.numeric(bin_mid),
  n_overlap_pop = as.integer(ceu_within_bin$n_segments),
  n_overlap_allpops = as.integer(all_within_bin$n_segments),
  overlap_bp_pop = as.numeric(ceu_within_bin$weight),
  overlap_bp_allpops = as.numeric(all_within_bin$weight),
  V_pop_within = as.numeric(ceu_within_bin$mean),
  V_allpops_within = as.numeric(all_within_bin$mean),
  V_within = as.numeric(v_within),
  V_total_grouped = as.numeric(v_total),
  FST_tree_raw = as.numeric(fst_raw),
  FST_tree_clipped = as.numeric(fst_clipped)
)

readr::write_csv(fst_df, out_fst_local_csv)

cat("Saved local F_ST^Tree (CEU vs allpops):", out_fst_local_csv, "\n")
print(dplyr::slice_head(fst_df, n = 10))
```
