---
title: "chr2_20samples D1F (Post-Burn-in)"
author: "Codex"
date: "2026-02-16"
output: html_document
---

## Overview

This notebook computes per-tree **D1F** distances to the theoretical Kingman mean F-matrix for each population using the ARGweaver outputs in the new organized layout.

Pipeline:
1. Discover population/run folders under `raw_data/data/runs`.
2. Build a Kingman theoretical-mean proxy with `n=20`.
3. For each population, iterate over tree-sequence files (`*.tree`) and keep only post-burn-in MCMC samples (`sample > 90`).
4. Compute per-tree `D1F = sum(abs(F_tree - F_mean_kingman))`.
5. Save per-sample and per-population outputs under `raw_data/data/results`.

## Setup

```{r, message=FALSE}
library(ape)
library(phylodyn)
library(readr)
library(dplyr)
library(stringr)
library(tibble)
library(fmatrix)
library(progress)

options(stringsAsFactors = FALSE)
```

```{r}
project_dir <- normalizePath(".", winslash = "/", mustWork = TRUE)
data_root <- file.path(project_dir, "raw_data", "data")
runs_root <- file.path(data_root, "runs")
results_root <- file.path(data_root, "results")

stopifnot(dir.exists(runs_root))
if (!dir.exists(results_root)) dir.create(results_root, recursive = TRUE)

# Problem-specific settings
target_f_dim <- 39L
n_kingman <- target_f_dim + 1L  # fmatrix::kingman_m(n) returns (n-1)x(n-1)
burn_in_max <- 90L

run_dirs <- list.dirs(runs_root, recursive = FALSE, full.names = TRUE)
run_dirs <- run_dirs[grepl("^argweaver_", basename(run_dirs))]

run_table <- tibble(
  run_dir = run_dirs,
  run_name = basename(run_dirs),
  run_tag = sub("^argweaver_", "", basename(run_dirs)),
  pop = str_match(sub("^argweaver_", "", basename(run_dirs)), "_([A-Z]{3})20$")[, 2]
) %>%
  filter(!is.na(pop)) %>%
  arrange(pop, run_name)

if (nrow(run_table) == 0) {
  stop("No run directories matching pattern argweaver_*_<POP>20 were found in raw_data/data/runs.")
}

target_pops <- unique(run_table$pop)

cat("Discovered populations:", paste(target_pops, collapse = ", "), "\n")
cat("Using burn-in cutoff: sample <=", burn_in_max, "excluded\n")
```

## Kingman `n=20` theoretical-mean proxy F-matrix

```{r}
out_mean_rds <- file.path(results_root, sprintf("F_mean_kingman_n%d_proxy.rds", n_kingman))
F_mean_kingman <- fmatrix::kingman_m(n = n_kingman)

if (!all(dim(F_mean_kingman) == c(target_f_dim, target_f_dim))) {
  stop(sprintf(
    "Expected Kingman matrix to be %dx%d but got %s.",
    target_f_dim, target_f_dim, paste(dim(F_mean_kingman), collapse = "x")
  ))
}

saveRDS(F_mean_kingman, out_mean_rds)

cat("Saved Kingman mean proxy F-matrix:", out_mean_rds, "\n")
```

## Compute per-tree D1F distances (post-burn-in only)

```{r}
dist_d1 <- function(F1, F2) {
  sum(abs(F1 - F2))
}

conform_fmat <- function(Fm, target_dim = 39L) {
  d <- dim(Fm)
  if (any(d < target_dim)) {
    stop(sprintf(
      "F-matrix is smaller than %dx%d (%s).",
      target_dim, target_dim, paste(d, collapse = "x")
    ))
  }
  if (any(d > target_dim)) {
    Fm <- Fm[seq_len(target_dim), seq_len(target_dim), drop = FALSE]
  }
  Fm
}

process_run <- function(run_dir, run_tag, pop) {
  trees_dir <- file.path(run_dir, "trees")
  breaks_dir <- file.path(results_root, run_tag, "breakpoints")
  out_dir <- file.path(results_root, run_tag, "d1f_postburnin")
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  stopifnot(dir.exists(trees_dir))
  stopifnot(dir.exists(breaks_dir))

  tree_files <- list.files(trees_dir, pattern = "_sample[0-9]+\\.tree$", full.names = TRUE)
  if (length(tree_files) == 0) {
    warning(sprintf("[%s] No .tree files found in %s", pop, trees_dir))
    return(tibble())
  }

  run_manifest <- tibble(
    tree_path = tree_files,
    stem = tools::file_path_sans_ext(basename(tree_files)),
    sample = as.integer(str_match(basename(tree_files), "sample([0-9]+)\\.tree$")[, 2]),
    breaks_path = file.path(breaks_dir, paste0(tools::file_path_sans_ext(basename(tree_files)), "_breaks.csv"))
  ) %>%
    filter(!is.na(sample)) %>%
    filter(sample > burn_in_max) %>%
    arrange(sample)

  if (nrow(run_manifest) == 0) {
    warning(sprintf("[%s] No post-burn-in .tree files (sample > %d) in %s", pop, burn_in_max, trees_dir))
    return(tibble())
  }

  missing_breaks <- !file.exists(run_manifest$breaks_path)
  if (any(missing_breaks)) {
    warning(sprintf(
      "[%s] Missing breaks files for %d sample(s). They will be skipped.",
      pop, sum(missing_breaks)
    ))
    run_manifest <- run_manifest[!missing_breaks, , drop = FALSE]
  }

  if (nrow(run_manifest) == 0) {
    warning(sprintf("[%s] All post-burn-in samples were skipped due to missing breaks files.", pop))
    return(tibble())
  }

  cat("Population:", pop, "| Run:", run_tag, "\n")
  cat("Post-burn-in samples to process:", nrow(run_manifest), "\n")

  all_results <- vector("list", nrow(run_manifest))

  for (j in seq_len(nrow(run_manifest))) {
    tree_path <- run_manifest$tree_path[j]
    breaks_path <- run_manifest$breaks_path[j]
    sample <- run_manifest$sample[j]
    stem <- run_manifest$stem[j]

    trees <- ape::read.tree(tree_path)
    if (inherits(trees, "phylo")) {
      trees <- structure(list(trees), class = "multiPhylo")
    }

    breaks_df <- readr::read_csv(breaks_path, show_col_types = FALSE) %>%
      arrange(tree_index)

    n_trees <- length(trees)
    pb <- progress::progress_bar$new(
      format = sprintf("Computing D1F (%s sample %d) [:bar] :current/:total (:percent) eta=:eta", pop, sample),
      total = n_trees,
      clear = FALSE,
      width = 60
    )

    d1f_vals <- vapply(seq_len(n_trees), function(i) {
      pb$tick()
      Fm <- phylodyn::gen_Fmat(trees[[i]], tol = 8L)
      Fm <- conform_fmat(Fm, target_dim = target_f_dim)
      dist_d1(Fm, F_mean_kingman)
    }, numeric(1))

    n_use <- min(length(d1f_vals), nrow(breaks_df))
    if (length(d1f_vals) != nrow(breaks_df)) {
      warning(sprintf(
        "Length mismatch for %s (trees=%d, breaks=%d). Truncating to %d.",
        stem, length(d1f_vals), nrow(breaks_df), n_use
      ))
    }

    res <- tibble(
      population = pop,
      run_tag = run_tag,
      sample = sample,
      file = breaks_df$file[seq_len(n_use)],
      tree_index = breaks_df$tree_index[seq_len(n_use)],
      left = breaks_df$left[seq_len(n_use)],
      right = breaks_df$right[seq_len(n_use)],
      mid = breaks_df$mid[seq_len(n_use)],
      d1f_distance = d1f_vals[seq_len(n_use)]
    )

    out_sample_csv <- file.path(out_dir, sprintf("%s_d1f_per_tree_postburnin.csv", stem))
    readr::write_csv(res, out_sample_csv)
    cat("Saved:", out_sample_csv, "\n")

    all_results[[j]] <- res
  }

  dplyr::bind_rows(all_results)
}
```

```{r}
all_d1f <- purrr::pmap_dfr(
  list(run_table$run_dir, run_table$run_tag, run_table$pop),
  process_run
)

cat("\nTotal post-burn-in rows:", nrow(all_d1f), "\n")
```

## Save per-population aggregated outputs

```{r}
if (nrow(all_d1f) > 0) {
  by_pop <- split(all_d1f, all_d1f$population)
  out_pop_dir <- file.path(results_root, "d1f_postburnin_n20_by_population")
  if (!dir.exists(out_pop_dir)) dir.create(out_pop_dir, recursive = TRUE)

  for (pop in names(by_pop)) {
    out_csv <- file.path(out_pop_dir, sprintf("chr2_20samples_%s_d1f_per_tree_postburnin.csv", pop))
    readr::write_csv(by_pop[[pop]], out_csv)
    cat("Saved aggregated file:", out_csv, "\n")
  }
} else {
  warning("No post-burn-in D1F rows were produced.")
}
```

## Quick summary

```{r}
if (nrow(all_d1f) > 0) {
  summary_tbl <- all_d1f %>%
    group_by(population, sample) %>%
    summarise(
      n_trees = n(),
      mean_d1f = mean(d1f_distance, na.rm = TRUE),
      median_d1f = median(d1f_distance, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(population, sample)

  print(summary_tbl, n = 200)
}
```

## Empirical Neutral Reference (Weighted Mean F) and D1F

```{r}
# -----------------------------------------------------------------------------
# Empirical neutral reference construction (population-specific)
# -----------------------------------------------------------------------------

# Expected layout for neutral ARGweaver runs inferred from neutral SLiM data:
# raw_data/data/runs_neutral_slim_argweaver/
#   argweaver_<ANYTAG>_<POP>20[_neutral...]/trees/*.tree
# and matching breakpoints in:
# raw_data/data/results/<ANYTAG>_<POP>20[_neutral...]/breakpoints/*_breaks.csv

neutral_runs_root <- file.path(data_root, "runs_neutral_slim_argweaver")
if (!dir.exists(neutral_runs_root)) {
  stop("Missing neutral run root: ", neutral_runs_root,
       "\nCreate this directory with ARGweaver outputs inferred from neutral SLiM simulations.")
}

dist_d1 <- function(F1, F2) {
  sum(abs(F1 - F2))
}

break_coalescent_ties_local <- function(tr, eps = 1e-8) {
  n <- tr$Nnode + 1L
  depths <- ape::node.depth.edgelength(tr)
  ages <- max(depths) - depths
  int <- (n + 1):(n + tr$Nnode)
  a <- ages[int]

  o <- order(a)
  a_sorted <- a[o]
  for (i in 2:length(a_sorted)) {
    if (a_sorted[i] <= a_sorted[i - 1]) a_sorted[i] <- a_sorted[i - 1] + eps
  }
  a[o] <- a_sorted
  ages[int] <- a

  new_depths <- max(ages) - ages
  parent <- tr$edge[, 1]
  child <- tr$edge[, 2]
  bl <- pmax(new_depths[child] - new_depths[parent], eps)
  tr$edge.length <- bl
  tr
}

safe_gen_fmat <- function(tr, tol = 8L, target_dim = target_f_dim) {
  # Attempt 1: direct
  out <- tryCatch(
    {
      Fm <- phylodyn::gen_Fmat(tr, tol = tol)
      conform_fmat(Fm, target_dim = target_dim)
    },
    error = function(e) NULL
  )
  if (!is.null(out)) return(out)

  # Attempt 2: resolve multifurcations + ties
  tr2 <- tryCatch(ape::multi2di(tr), error = function(e) tr)
  tr2 <- break_coalescent_ties_local(tr2, eps = 1e-8)
  out <- tryCatch(
    {
      Fm <- phylodyn::gen_Fmat(tr2, tol = tol)
      conform_fmat(Fm, target_dim = target_dim)
    },
    error = function(e) NULL
  )
  out
}

compute_weighted_F_from_treefile <- function(tree_path, breaks_path, target_dim = target_f_dim) {
  trees <- ape::read.tree(tree_path)
  if (inherits(trees, "phylo")) {
    trees <- structure(list(trees), class = "multiPhylo")
  }

  breaks_df <- readr::read_csv(breaks_path, show_col_types = FALSE) %>%
    arrange(tree_index)

  n_trees <- length(trees)
  n_use <- min(n_trees, nrow(breaks_df))
  if (n_use == 0) return(NULL)

  widths <- breaks_df$right[seq_len(n_use)] - breaks_df$left[seq_len(n_use)]
  if (sum(widths) <= 0) return(NULL)

  F_list <- vector("list", n_use)
  for (i in seq_len(n_use)) {
    Fm <- safe_gen_fmat(trees[[i]], tol = 8L, target_dim = target_dim)
    if (is.null(Fm)) next
    F_list[[i]] <- Fm
  }

  keep <- which(!vapply(F_list, is.null, logical(1)))
  if (length(keep) == 0) return(NULL)
  F_list <- F_list[keep]
  widths <- widths[keep]

  F_sum <- matrix(0, nrow = target_dim, ncol = target_dim)
  for (i in seq_along(F_list)) {
    F_sum <- F_sum + widths[i] * F_list[[i]]
  }

  F_sum / sum(widths)
}

build_empirical_reference_for_pop <- function(pop, burn_cutoff = burn_in_max) {
  neutral_run_dirs <- list.dirs(neutral_runs_root, recursive = FALSE, full.names = TRUE)
  neutral_run_dirs <- neutral_run_dirs[grepl("^argweaver_", basename(neutral_run_dirs))]
  neutral_run_dirs <- neutral_run_dirs[
    grepl(pop, basename(neutral_run_dirs), ignore.case = TRUE) &
      grepl("neutral", basename(neutral_run_dirs), ignore.case = TRUE)
  ]

  if (length(neutral_run_dirs) == 0) {
    stop("No neutral run dirs found for population ", pop,
         " under ", neutral_runs_root,
         " (expected names containing both population code and 'neutral').")
  }

  pb_runs <- progress::progress_bar$new(
    format = sprintf("Empirical ref (%s) runs [:bar] :current/:total (:percent) eta=:eta", pop),
    total = length(neutral_run_dirs),
    clear = FALSE,
    width = 70
  )

  weighted_list <- list()
  for (run_dir in neutral_run_dirs) {
    pb_runs$tick()
    run_tag <- sub("^argweaver_", "", basename(run_dir))
    trees_dir <- file.path(run_dir, "trees")
    breaks_dir <- file.path(results_root, run_tag, "breakpoints")

    if (!dir.exists(trees_dir) || !dir.exists(breaks_dir)) {
      warning("Skipping neutral run (missing trees or breakpoints): ", run_dir)
      next
    }

    tree_files <- list.files(trees_dir, pattern = "_sample[0-9]+\\.tree$", full.names = TRUE)
    if (length(tree_files) == 0) next

    manifest <- tibble(
      tree_path = tree_files,
      stem = tools::file_path_sans_ext(basename(tree_files)),
      sample = as.integer(stringr::str_match(basename(tree_files), "sample([0-9]+)\\.tree$")[, 2]),
      breaks_path = file.path(
        breaks_dir,
        paste0(tools::file_path_sans_ext(basename(tree_files)), "_breaks.csv")
      )
    ) %>%
      filter(!is.na(sample)) %>%
      filter(sample > burn_cutoff) %>%
      arrange(sample)

    if (nrow(manifest) == 0) next

    pb_samples <- progress::progress_bar$new(
      format = sprintf("Empirical ref (%s %s) samples [:bar] :current/:total (:percent) eta=:eta",
                       pop, run_tag),
      total = nrow(manifest),
      clear = FALSE,
      width = 70
    )

    for (i in seq_len(nrow(manifest))) {
      pb_samples$tick()
      cat(sprintf("Empirical ref input: %s sample=%d\n", run_tag, manifest$sample[i]))
      if (!file.exists(manifest$breaks_path[i])) next
      Fw <- compute_weighted_F_from_treefile(
        tree_path = manifest$tree_path[i],
        breaks_path = manifest$breaks_path[i],
        target_dim = target_f_dim
      )
      if (!is.null(Fw)) {
        weighted_list[[length(weighted_list) + 1L]] <- Fw
      }
    }
  }

  if (length(weighted_list) == 0) {
    stop("No usable neutral weighted F-matrices found for population ", pop)
  }

  F_ref <- Reduce(`+`, weighted_list) / length(weighted_list)

  out_ref <- file.path(results_root, sprintf("F_empirical_weighted_neutral_%s_n20.rds", pop))
  saveRDS(F_ref, out_ref)
  cat("Saved empirical weighted neutral reference:", out_ref,
      "| n_weighted_samples=", length(weighted_list), "\n")

  F_ref
}

F_empirical_by_pop <- setNames(vector("list", length(target_pops)), target_pops)
for (pop in target_pops) {
  F_empirical_by_pop[[pop]] <- build_empirical_reference_for_pop(pop)
}
```

```{r}
# -----------------------------------------------------------------------------
# Recompute post-burn-in D1F to empirical neutral reference (population-specific)
# -----------------------------------------------------------------------------

process_run_empirical <- function(run_dir, run_tag, pop) {
  trees_dir <- file.path(run_dir, "trees")
  breaks_dir <- file.path(results_root, run_tag, "breakpoints")
  out_dir <- file.path(results_root, run_tag, "d1f_postburnin_empirical_neutral")
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  stopifnot(dir.exists(trees_dir))
  stopifnot(dir.exists(breaks_dir))

  F_ref <- F_empirical_by_pop[[pop]]
  if (is.null(F_ref)) stop("Missing empirical reference for pop: ", pop)

  tree_files <- list.files(trees_dir, pattern = "_sample[0-9]+\\.tree$", full.names = TRUE)
  if (length(tree_files) == 0) {
    warning(sprintf("[%s] No .tree files found in %s", pop, trees_dir))
    return(tibble())
  }

  run_manifest <- tibble(
    tree_path = tree_files,
    stem = tools::file_path_sans_ext(basename(tree_files)),
    sample = as.integer(str_match(basename(tree_files), "sample([0-9]+)\\.tree$")[, 2]),
    breaks_path = file.path(breaks_dir, paste0(tools::file_path_sans_ext(basename(tree_files)), "_breaks.csv"))
  ) %>%
    filter(!is.na(sample)) %>%
    filter(sample > burn_in_max) %>%
    arrange(sample)

  if (nrow(run_manifest) == 0) {
    warning(sprintf("[%s] No post-burn-in .tree files (sample > %d) in %s", pop, burn_in_max, trees_dir))
    return(tibble())
  }

  missing_breaks <- !file.exists(run_manifest$breaks_path)
  if (any(missing_breaks)) {
    warning(sprintf(
      "[%s] Missing breaks files for %d sample(s). They will be skipped.",
      pop, sum(missing_breaks)
    ))
    run_manifest <- run_manifest[!missing_breaks, , drop = FALSE]
  }

  if (nrow(run_manifest) == 0) {
    warning(sprintf("[%s] All post-burn-in samples were skipped due to missing breaks files.", pop))
    return(tibble())
  }

  all_results <- vector("list", nrow(run_manifest))

  pb_samples <- progress::progress_bar$new(
    format = sprintf("D1F empirical (%s %s) samples [:bar] :current/:total (:percent) eta=:eta",
                     pop, run_tag),
    total = nrow(run_manifest),
    clear = FALSE,
    width = 70
  )

  for (j in seq_len(nrow(run_manifest))) {
    pb_samples$tick()
    tree_path <- run_manifest$tree_path[j]
    breaks_path <- run_manifest$breaks_path[j]
    sample <- run_manifest$sample[j]
    stem <- run_manifest$stem[j]

    trees <- ape::read.tree(tree_path)
    if (inherits(trees, "phylo")) {
      trees <- structure(list(trees), class = "multiPhylo")
    }

    breaks_df <- readr::read_csv(breaks_path, show_col_types = FALSE) %>%
      arrange(tree_index)

    n_trees <- length(trees)
    pb_trees <- progress::progress_bar$new(
      format = sprintf("D1F empirical (%s sample %d) trees [:bar] :current/:total (:percent) eta=:eta",
                       pop, sample),
      total = n_trees,
      clear = FALSE,
      width = 70
    )

    d1f_vals <- rep(NA_real_, n_trees)
    for (i in seq_len(n_trees)) {
      pb_trees$tick()
      Fm <- safe_gen_fmat(trees[[i]], tol = 8L, target_dim = target_f_dim)
      if (is.null(Fm)) next
      d1f_vals[i] <- dist_d1(Fm, F_ref)
    }

    n_use <- min(length(d1f_vals), nrow(breaks_df))
    if (length(d1f_vals) != nrow(breaks_df)) {
      warning(sprintf(
        "Length mismatch for %s (trees=%d, breaks=%d). Truncating to %d.",
        stem, length(d1f_vals), nrow(breaks_df), n_use
      ))
    }

    if (all(is.na(d1f_vals[seq_len(n_use)]))) {
      warning(sprintf("All trees failed F-matrix generation for %s; skipping.", stem))
      next
    }

    res <- tibble(
      population = pop,
      run_tag = run_tag,
      sample = sample,
      file = breaks_df$file[seq_len(n_use)],
      tree_index = breaks_df$tree_index[seq_len(n_use)],
      left = breaks_df$left[seq_len(n_use)],
      right = breaks_df$right[seq_len(n_use)],
      mid = breaks_df$mid[seq_len(n_use)],
      d1f_distance = d1f_vals[seq_len(n_use)]
    )

    out_sample_csv <- file.path(out_dir, sprintf("%s_d1f_per_tree_postburnin_empirical_neutral.csv", stem))
    readr::write_csv(res, out_sample_csv)
    cat("Saved:", out_sample_csv, "\n")

    all_results[[j]] <- res
  }

  dplyr::bind_rows(all_results)
}

all_d1f_empirical <- purrr::pmap_dfr(
  list(run_table$run_dir, run_table$run_tag, run_table$pop),
  process_run_empirical
)

cat("\nTotal post-burn-in rows (empirical neutral reference):", nrow(all_d1f_empirical), "\n")
```

```{r}
# Save per-population aggregated outputs for empirical-reference D1F
if (nrow(all_d1f_empirical) > 0) {
  by_pop_emp <- split(all_d1f_empirical, all_d1f_empirical$population)
  out_pop_dir_emp <- file.path(results_root, "d1f_postburnin_empirical_n20_by_population")
  if (!dir.exists(out_pop_dir_emp)) dir.create(out_pop_dir_emp, recursive = TRUE)

  for (pop in names(by_pop_emp)) {
    out_csv <- file.path(out_pop_dir_emp, sprintf("chr2_20samples_%s_d1f_per_tree_postburnin_empirical_neutral.csv", pop))
    readr::write_csv(by_pop_emp[[pop]], out_csv)
    cat("Saved aggregated empirical-reference file:", out_csv, "\n")
  }

  summary_tbl_emp <- all_d1f_empirical %>%
    group_by(population, sample) %>%
    summarise(
      n_trees = n(),
      mean_d1f = mean(d1f_distance, na.rm = TRUE),
      median_d1f = median(d1f_distance, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(population, sample)

  print(summary_tbl_emp, n = 200)
} else {
  warning("No empirical-reference D1F rows were produced.")
}
```
